---
title: "Boosting Ensemble Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    caTools,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    ranger,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    xgboost,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Introduction**

Boosting is a tree-based ensemble method designed to improve predictive performance by combining multiple weak learners. In most classification applications, these weak learners are shallow decision trees that provide limited predictive accuracy individually but become substantially more powerful when combined. Unlike random forests, which train many trees independently and aggregate their predictions through majority voting or probability averaging, boosting trains trees sequentially. Each new tree is constructed to improve on the errors made by the previous trees, allowing the model to focus progressively on observations that are more difficult to classify.

The boosting process begins with an initial weak model fit to the training data. Observations that are incorrectly classified or poorly predicted are given greater emphasis in subsequent iterations. As additional trees are added, each model attempts to correct the remaining error from the existing ensemble. The final prediction is based on the combined contribution of all trees, where trees that reduce prediction error more effectively have greater influence on the final classification. This sequential learning process can improve model accuracy, but it also requires careful tuning because boosted models may overfit if too many trees are added or if individual trees are too complex.

Boosting differs from random forests in both construction and modeling objective. Random forests reduce prediction variance by averaging many independently grown trees, whereas boosting sequentially reduces error by emphasizing difficult cases. Random forest trees are often grown deeply and decorrelated through bootstrap sampling and random feature selection. Boosted trees are typically shallower and are added one at a time, with each tree contributing an incremental improvement. Consequently, boosting can provide strong predictive performance but may be slower to train and more sensitive to hyperparameter choices than random forests.

Several boosting algorithms are available for classification problems. AdaBoost increases the weight of misclassified observations so that subsequent weak learners focus more heavily on those cases. Gradient boosting methods, including GBM, XGBoost, LightGBM, and CatBoost, improve the model by fitting new trees to the gradient of the loss function. LogitBoost is another classification-focused boosting method that updates predictions using a log-odds formulation. In this project, gradient boosting was evaluated using the `xgboost` package in R because it provides an efficient and flexible implementation of boosted decision trees.

The performance of XGBoost depends on several tuning parameters. The number of boosting rounds, `nrounds`, controls how many trees are added to the ensemble. The `max_depth` parameter controls the complexity of each tree, while `eta` determines the learning rate, or the size of each boosting update. Smaller learning rates often require more boosting rounds but can improve generalization. The `gamma` parameter specifies the minimum loss reduction required to create a split. The `subsample` and `colsample_bytree` parameters introduce randomness by sampling rows and predictors during training, which can reduce overfitting. Finally, `min_child_weight` controls the minimum amount of information required in a child node and can prevent overly specific splits.

In this project, boosted tree models were evaluated for binary classification of loan default. Because loan default is an imbalanced classification problem, XGBoost was used to estimate the probability that a borrower would default, and model performance was assessed using F1 score as the primary metric. Additional metrics, including accuracy, precision, recall, balanced accuracy, and area under the receiver operating characteristic curve, were used to compare boosted models with decision trees, random forests, logistic regression models, and neural networks.

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS("train.rds")
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS("train_under.rds")
```

## *Holdout Dataset*

```{r}
hold <- readRDS("holdout.rds")
```

# **Preview the Data**

## *Training Dataset*

```{r}
skimr::skim(train)
```

## *Undersampled Training Dataset*

```{r}
skimr::skim(train_under)
```

## *Holdout Dataset*

```{r}
skimr::skim(hold)
```

# **Create `evaluate_classifier()`**

```{r}
evaluate_classifier <- function(actual, predicted, probability_yes, positive_class = "Yes") {
  actual <- factor(actual, levels = c("Yes", "No"))
  predicted <- factor(predicted, levels = c("Yes", "No"))

  cm <- caret::confusionMatrix(
    data = predicted,
    reference = actual,
    positive = positive_class
  )

  roc_obj <- pROC::roc(
    response = actual,
    predictor = probability_yes,
    levels = c("No", "Yes"),
    direction = "<",
    quiet = TRUE
  )

  precision <- cm$byClass["Precision"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall / (precision + recall)
  )

  tibble::tibble(
    accuracy = as.numeric(cm$overall["Accuracy"]),
    balanced_accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    precision = as.numeric(precision),
    sensitivity_recall = as.numeric(recall),
    specificity = as.numeric(cm$byClass["Specificity"]),
    f1_score = as.numeric(f1),
    auc = as.numeric(pROC::auc(roc_obj))
  )
}
```

# **Shallow XGBoost Models**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_shallow_params <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  max_depth = 3,
  eta = 0.08,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 3
)

xgb_shallow_nrounds <- 120
xgb_shallow_threshold <- 0.62
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_shallow <- xgboost::xgboost(
  data = data.matrix(train %>% select(-loan_default)),
  label = ifelse(train$loan_default == "Yes", 1, 0),
  params = xgb_shallow_params,
  nrounds = xgb_shallow_nrounds,
  verbose = 0
)
```


### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_shallow_prob <- predict(
  object = xgb_shallow,
  newdata = data.matrix(hold %>% select(-loan_default))
)
```

#### Predict Class

```{r}
xgb_shallow_class <- factor(
  ifelse(xgb_shallow_prob >= xgb_shallow_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_shallow_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_shallow_class,
  probability_yes = as.numeric(xgb_shallow_prob)
)
```

#### Preview Table

```{r}
head(xgb_shallow_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_shallow_metrics <- evaluate_classifier(
  actual = xgb_shallow_predictions$actual,
  predicted = xgb_shallow_predictions$predicted,
  probability_yes = xgb_shallow_predictions$probability_yes
)

xgb_shallow_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_shallow, file.path(project_dir, "xgb_shallow_model.rds"))
```

#### Save Results

```{r}
xgb_shallow_results <- xgb_shallow_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_shallow",
    sample = "Full",
    threshold = xgb_shallow_threshold,
    nrounds = xgb_shallow_nrounds,
    max_depth = xgb_shallow_params$max_depth,
    eta = xgb_shallow_params$eta,
    subsample = xgb_shallow_params$subsample,
    colsample_bytree = xgb_shallow_params$colsample_bytree,
    min_child_weight = xgb_shallow_params$min_child_weight
  )

saveRDS(xgb_shallow_results, file.path(project_dir, "xgb_shallow_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_shallow_us_threshold <- 0.64
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_shallow_us <- xgboost::xgboost(
  data = data.matrix(train_under %>% select(-loan_default)),
  label = ifelse(train_under$loan_default == "Yes", 1, 0),
  params = xgb_shallow_params,
  nrounds = xgb_shallow_nrounds,
  verbose = 0
)
```


### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_shallow_us_prob <- predict(xgb_shallow_us, newdata = data.matrix(hold %>% select(-loan_default)))
```

#### Predict Class

```{r}
xgb_shallow_us_class <- factor(
  ifelse(xgb_shallow_us_prob >= xgb_shallow_us_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_shallow_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_shallow_us_class,
  probability_yes = as.numeric(xgb_shallow_us_prob)
)
```

#### Preview Table

```{r}
head(xgb_shallow_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_shallow_us_metrics <- evaluate_classifier(
  actual = xgb_shallow_us_predictions$actual,
  predicted = xgb_shallow_us_predictions$predicted,
  probability_yes = xgb_shallow_us_predictions$probability_yes
)

xgb_shallow_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_shallow_us, file.path(project_dir, "xgb_shallow_us_model.rds"))
```

#### Save Results

```{r}

xgb_shallow_us_results <- xgb_shallow_us_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_shallow",
    sample = "Undersampled",
    threshold = xgb_shallow_us_threshold,
    nrounds = xgb_shallow_nrounds,
    max_depth = xgb_shallow_params$max_depth,
    eta = xgb_shallow_params$eta,
    subsample = xgb_shallow_params$subsample,
    colsample_bytree = xgb_shallow_params$colsample_bytree,
    min_child_weight = xgb_shallow_params$min_child_weight
  )

saveRDS(xgb_shallow_us_results, file.path(project_dir, "xgb_shallow_us_results.rds"))
```

# **Medium XGBoost Models**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_medium_params <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  max_depth = 5,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 3
)

xgb_medium_nrounds <- 180
xgb_medium_threshold <- 0.65
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_medium <- xgboost::xgboost(
  data = data.matrix(train %>% select(-loan_default)),
  label = ifelse(train$loan_default == "Yes", 1, 0),
  params = xgb_medium_params,
  nrounds = xgb_medium_nrounds,
  verbose = 0
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_medium_prob <- predict(
  object = xgb_medium,
  newdata = data.matrix(hold %>% select(-loan_default))
)
```

#### Predict Class

```{r}
xgb_medium_class <- factor(
  ifelse(xgb_medium_prob >= xgb_medium_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_medium_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_medium_class,
  probability_yes = as.numeric(xgb_medium_prob)
)
```

#### Preview Table

```{r}
head(xgb_medium_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_medium_metrics <- evaluate_classifier(
  actual = xgb_medium_predictions$actual,
  predicted = xgb_medium_predictions$predicted,
  probability_yes = xgb_medium_predictions$probability_yes
)

xgb_medium_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_medium, file.path(project_dir, "xgb_medium_model.rds"))
```

#### Save Results

```{r}
xgb_medium_results <- xgb_medium_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_medium",
    sample = "Full",
    threshold = xgb_medium_threshold,
    nrounds = xgb_medium_nrounds,
    max_depth = xgb_medium_params$max_depth,
    eta = xgb_medium_params$eta,
    subsample = xgb_medium_params$subsample,
    colsample_bytree = xgb_medium_params$colsample_bytree,
    min_child_weight = xgb_medium_params$min_child_weight
  )

saveRDS(xgb_medium_results, file.path(project_dir, "xgb_medium_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_medium_us_threshold <- 0.66
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_medium_us <- xgboost::xgboost(
  data = data.matrix(train_under %>% select(-loan_default)),
  label = ifelse(train_under$loan_default == "Yes", 1, 0),
  params = xgb_medium_params,
  nrounds = xgb_medium_nrounds,
  verbose = 0
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_medium_us_prob <- predict(
  object = xgb_medium_us,
  newdata = data.matrix(hold %>% select(-loan_default))
)
```

#### Predict Class

```{r}
xgb_medium_us_class <- factor(
  ifelse(xgb_medium_us_prob >= xgb_medium_us_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_medium_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_medium_us_class,
  probability_yes = as.numeric(xgb_medium_us_prob)
)
```

#### Preview Table

```{r}
head(xgb_medium_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_medium_us_metrics <- evaluate_classifier(
  actual = xgb_medium_us_predictions$actual,
  predicted = xgb_medium_us_predictions$predicted,
  probability_yes = xgb_medium_us_predictions$probability_yes
)

xgb_medium_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_medium_us, file.path(project_dir, "xgb_medium_us_model.rds"))
```

#### Save Results

```{r}
xgb_medium_us_results <- xgb_medium_us_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_medium",
    sample = "Undersampled",
    threshold = xgb_medium_us_threshold,
    nrounds = xgb_medium_nrounds,
    max_depth = xgb_medium_params$max_depth,
    eta = xgb_medium_params$eta,
    subsample = xgb_medium_params$subsample,
    colsample_bytree = xgb_medium_params$colsample_bytree,
    min_child_weight = xgb_medium_params$min_child_weight
  )

saveRDS(xgb_medium_us_results, file.path(project_dir, "xgb_medium_us_results.rds"))
```

# **Slow XGBoost Models**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_slow_params <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  max_depth = 4,
  eta = 0.03,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 3
)

xgb_slow_nrounds <- 260
xgb_slow_threshold <- 0.65
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_slow <- xgboost::xgboost(
  data = data.matrix(train %>% select(-loan_default)),
  label = ifelse(train$loan_default == "Yes", 1, 0),
  params = xgb_slow_params,
  nrounds = xgb_slow_nrounds,
  verbose = 0
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_slow_prob <- predict(
  object = xgb_slow,
  newdata = data.matrix(hold %>% select(-loan_default))
)
```

#### Predict Class

```{r}
xgb_slow_class <- factor(
  ifelse(xgb_slow_prob >= xgb_slow_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_slow_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_slow_class,
  probability_yes = as.numeric(xgb_slow_prob)
)
```

#### Preview Table

```{r}
head(xgb_slow_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_slow_metrics <- evaluate_classifier(
  actual = xgb_slow_predictions$actual,
  predicted = xgb_slow_predictions$predicted,
  probability_yes = xgb_slow_predictions$probability_yes
)

xgb_slow_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_slow, file.path(project_dir, "xgb_slow_model.rds"))
```

#### Save Results

```{r}
xgb_slow_results <- xgb_slow_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_slow",
    sample = "Full",
    threshold = xgb_slow_threshold,
    nrounds = xgb_slow_nrounds,
    max_depth = xgb_slow_params$max_depth,
    eta = xgb_slow_params$eta,
    subsample = xgb_slow_params$subsample,
    colsample_bytree = xgb_slow_params$colsample_bytree,
    min_child_weight = xgb_slow_params$min_child_weight
  )

saveRDS(xgb_slow_results, file.path(project_dir, "xgb_slow_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
xgb_slow_us_threshold <- 0.61
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

xgb_slow_us <- xgboost::xgboost(
  data = data.matrix(train_under %>% select(-loan_default)),
  label = ifelse(train_under$loan_default == "Yes", 1, 0),
  params = xgb_slow_params,
  nrounds = xgb_slow_nrounds,
  verbose = 0
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
xgb_slow_us_prob <- predict(
  object = xgb_slow_us,
  newdata = data.matrix(hold %>% select(-loan_default))
)
```

#### Predict Class

```{r}
xgb_slow_us_class <- factor(
  ifelse(xgb_slow_us_prob >= xgb_slow_us_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
xgb_slow_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = xgb_slow_us_class,
  probability_yes = as.numeric(xgb_slow_us_prob)
)
```

#### Preview Table

```{r}
head(xgb_slow_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
xgb_slow_us_metrics <- evaluate_classifier(
  actual = xgb_slow_us_predictions$actual,
  predicted = xgb_slow_us_predictions$predicted,
  probability_yes = xgb_slow_us_predictions$probability_yes
)

xgb_slow_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(xgb_slow_us, file.path(project_dir, "xgb_slow_us_model.rds"))
```

#### Save Results

```{r}
xgb_slow_us_results <- xgb_slow_us_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "xgboost",
    model_variant = "xgb_slow",
    sample = "Undersampled",
    threshold = xgb_slow_us_threshold,
    nrounds = xgb_slow_nrounds,
    max_depth = xgb_slow_params$max_depth,
    eta = xgb_slow_params$eta,
    subsample = xgb_slow_params$subsample,
    colsample_bytree = xgb_slow_params$colsample_bytree,
    min_child_weight = xgb_slow_params$min_child_weight
  )

saveRDS(xgb_slow_us_results, file.path(project_dir, "xgb_slow_us_results.rds"))
```

# **LogitBoost Models - 50 Iterations**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
logitboost_50_niter <- 50
logitboost_50_threshold <- 0.12
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

logitboost_50 <- caTools::LogitBoost(
  data.matrix(train %>% select(-loan_default)),
  ifelse(train$loan_default == "Yes", 1, 0),
  nIter = logitboost_50_niter
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
logitboost_50_prob <- predict(
  object = logitboost_50,
  data.matrix(hold %>% select(-loan_default)),
  type = "raw"
)

logitboost_50_prob <- as.numeric(logitboost_50_prob[, "1"])
```

#### Predict Class

```{r}
logitboost_50_class <- factor(
  ifelse(logitboost_50_prob >= logitboost_50_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
logitboost_50_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = logitboost_50_class,
  probability_yes = as.numeric(logitboost_50_prob)
)
```

#### Preview Table

```{r}
head(logitboost_50_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
logitboost_50_metrics <- evaluate_classifier(
  actual = logitboost_50_predictions$actual,
  predicted = logitboost_50_predictions$predicted,
  probability_yes = logitboost_50_predictions$probability_yes
)

logitboost_50_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(logitboost_50, file.path(project_dir, "logitboost_50_model.rds"))
```

#### Save Results

```{r}
logitboost_50_results <- logitboost_50_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "caTools",
    model_variant = "logitboost_50",
    sample = "Full",
    threshold = logitboost_50_threshold,
    nIter = logitboost_50_niter
  )

saveRDS(logitboost_50_results, file.path(project_dir, "logitboost_50_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
logitboost_50_us_threshold <- 0.51
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

logitboost_50_us <- caTools::LogitBoost(
  data.matrix(train_under %>% select(-loan_default)),
  ifelse(train_under$loan_default == "Yes", 1, 0),
  nIter = logitboost_50_niter
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
logitboost_50_us_prob <- predict(
  object = logitboost_50_us,
  data.matrix(hold %>% select(-loan_default)),
  type = "raw"
)

logitboost_50_us_prob <- as.numeric(logitboost_50_us_prob[, "1"])
```

#### Predict Class

```{r}
logitboost_50_us_class <- factor(
  ifelse(logitboost_50_us_prob >= logitboost_50_us_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
logitboost_50_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = logitboost_50_us_class,
  probability_yes = as.numeric(logitboost_50_us_prob)
)
```

#### Preview Table

```{r}
head(logitboost_50_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
logitboost_50_us_metrics <- evaluate_classifier(
  actual = logitboost_50_us_predictions$actual,
  predicted = logitboost_50_us_predictions$predicted,
  probability_yes = logitboost_50_us_predictions$probability_yes
)

logitboost_50_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(logitboost_50_us, file.path(project_dir, "logitboost_50_us_model.rds"))
```

#### Save Results

```{r}
logitboost_50_us_results <- logitboost_50_us_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "caTools",
    model_variant = "logitboost_50",
    sample = "Undersampled",
    threshold = logitboost_50_us_threshold,
    nIter = logitboost_50_niter
  )

saveRDS(logitboost_50_us_results, file.path(project_dir, "logitboost_50_us_results.rds"))
```

# **LogitBoost Models - 100 Iterations**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
logitboost_100_niter <- 100
logitboost_100_threshold <- 0.12
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

logitboost_100 <- caTools::LogitBoost(
  data.matrix(train %>% select(-loan_default)),
  ifelse(train$loan_default == "Yes", 1, 0),
  nIter = logitboost_100_niter
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
logitboost_100_prob <- predict(
  object = logitboost_100,
  data.matrix(hold %>% select(-loan_default)),
  type = "raw"
)

logitboost_100_prob <- as.numeric(logitboost_100_prob[, "1"])
```

#### Predict Class

```{r}
logitboost_100_class <- factor(
  ifelse(logitboost_100_prob >= logitboost_100_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
logitboost_100_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = logitboost_100_class,
  probability_yes = as.numeric(logitboost_100_prob)
)
```

#### Preview Table

```{r}
head(logitboost_100_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
logitboost_100_metrics <- evaluate_classifier(
  actual = logitboost_100_predictions$actual,
  predicted = logitboost_100_predictions$predicted,
  probability_yes = logitboost_100_predictions$probability_yes
)

logitboost_100_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(logitboost_100, file.path(project_dir, "logitboost_100_model.rds"))
```

#### Save Results

```{r}
logitboost_100_results <- logitboost_100_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "caTools",
    model_variant = "logitboost_100",
    sample = "Full",
    threshold = logitboost_100_threshold,
    nIter = logitboost_100_niter
  )

saveRDS(logitboost_100_results, file.path(project_dir, "logitboost_100_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
logitboost_100_us_threshold <- 0.51
```

### Build Boosted Ensemble
____________________________________________________________________________________________________________

```{r}
set.seed(123)

logitboost_100_us <- caTools::LogitBoost(
  data.matrix(train_under %>% select(-loan_default)),
  ifelse(train_under$loan_default == "Yes", 1, 0),
  nIter = logitboost_100_niter
)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
logitboost_100_us_prob <- predict(
  object = logitboost_100_us,
  data.matrix(hold %>% select(-loan_default)),
  type = "raw"
)

logitboost_100_us_prob <- as.numeric(logitboost_100_us_prob[, "1"])
```

#### Predict Class

```{r}
logitboost_100_us_class <- factor(
  ifelse(logitboost_100_us_prob >= logitboost_100_us_threshold, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
logitboost_100_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = logitboost_100_us_class,
  probability_yes = as.numeric(logitboost_100_us_prob)
)
```

#### Preview Table

```{r}
head(logitboost_100_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
logitboost_100_us_metrics <- evaluate_classifier(
  actual = logitboost_100_us_predictions$actual,
  predicted = logitboost_100_us_predictions$predicted,
  probability_yes = logitboost_100_us_predictions$probability_yes
)

logitboost_100_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(logitboost_100_us, file.path(project_dir, "logitboost_100_us_model.rds"))
```

#### Save Results

```{r}
logitboost_100_us_results <- logitboost_100_us_metrics %>%
  mutate(
    model_type = "Boosted Tree Ensemble",
    package = "caTools",
    model_variant = "logitboost_100",
    sample = "Undersampled",
    threshold = logitboost_100_us_threshold,
    nIter = logitboost_100_niter
  )

saveRDS(logitboost_100_us_results, file.path(project_dir, "logitboost_100_us_results.rds"))
```