---
title: "Logistic Regression Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS("train.rds")
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS("train_under.rds")
```

## *Holdout Dataset*

```{r}
hold <- readRDS("holdout.rds")
```

# **Handling Target Variable**

## *Training Dataset*

```{r}
train_logit <- train %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

## *Undersampled Training Dataset*

```{r}
train_under_logit <- train %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

## *Holdout Dataset*

```{r}
hold_logit <- hold %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

# **Skim the Data**

```{r}
skimr::skim(train)
```

```{r}
skimr::skim(train_under)
```

```{r}
skimr::skim(hold)
```

# **Modeling**

## *Explanation of the Parameters*

### `train()` Parameters
____________________________________________________________________________________________________________

The `carat` package contains the function `train()` that sets up a grid of tuning parameters for a number of classification and regression routines, fits each model, and calculates a resampling based performance measure. This function fits predictive models over different tuning parameters. The different tuning parameters are described in **Table 1**. These parameters were utilized to perform logistic regression. The particular parameter that enables logistic regression in the `train()` function is `method`. Setting `method` to `glmnet` runs a logistic regression model.

**Table 1.** A tabulation of the `train()` function's parameters and corresponding descriptions of the parameters.

| Parameter | Description |
|:---|:---|
| `x` | Predictor data. For the default method, rows are samples and columns are features. This can be a matrix, data frame, sparse matrix, or similar object, but columns must be named. If using the recipe method, `x` should be an unprepared `recipes::recipe()` object that defines the outcome, predictors, and preprocessing steps. |
| `...` | Additional arguments passed to the underlying classification or regression routine. Tuning parameters should not be passed here because they belong in `tuneGrid` or `tuneLength`. |
| `y` | Outcome vector for each sample. This should be numeric for regression or a factor for classification. |
| `method` | String specifying the model type to train, such as `"glmnet"`, `"rpart"`, `"ranger"`, or `"nnet"`. Available model names can be checked with `names(getModelInfo())`. |
| `preProcess` | Character vector specifying preprocessing steps for predictor data, such as `"center"`, `"scale"`, `"range"`, `"medianImpute"`, `"pca"`, or `"YeoJohnson"`. This applies only when `x` is a matrix or data frame, not when using a recipe. |
| `weights` | Numeric vector of case weights. This only affects models that support weighted observations. |
| `metric` | Summary metric used to select the optimal model. Common defaults are `"RMSE"` and `"Rsquared"` for regression or `"Accuracy"` and `"Kappa"` for classification. If using a custom `summaryFunction`, `metric` should match one of the returned metric names. |
| `maximize` | Logical value indicating whether the selected metric should be maximized or minimized. |
| `trControl` | A `trainControl()` object that defines resampling, class probabilities, saved predictions, summary functions, sampling, seeds, and parallel behavior. |
| `tuneGrid` | Data frame of tuning parameter combinations. Column names must match the model’s tuning parameters, which can be checked with `modelLookup()` or `getModelInfo()`. |
| `tuneLength` | Integer controlling the size or granularity of the automatically generated tuning grid. If random search is used, this is the maximum number of tuning combinations. |
| `form` | Formula specifying the model structure, usually in the form `outcome ~ predictor1 + predictor2 + ...`. |
| `data` | Data frame containing the variables referenced in the formula or recipe. |
| `subset` | Index vector specifying which rows should be used in the training sample. |
| `na.action` | Function specifying what to do when missing values are found. The default is to fail; alternatives such as `na.omit` remove incomplete cases. |
| `contrasts` | List specifying contrasts for factor variables used in the model formula. |

Not all of the parameters described in **Table 1** will be utilized.

### `trainControl()` Parameters
____________________________________________________________________________________________________________

An explanation of the parameters in the `trainControl` object are described in **Table 2**.

**Table 2.** A tabulation of the parameters in `trainControl()` and their corresponding descriptions.

| Parameter | Description |
|:---|:---|
| `method` | The resampling method, such as `"boot"`, `"cv"`, `"repeatedcv"`, `"LOOCV"`, `"LGOCV"`, `"none"`, `"oob"`, `"timeslice"`, `"adaptive_cv"`, `"adaptive_boot"`, or `"adaptive_LGOCV"`. |
| `number` | The number of folds or resampling iterations. |
| `repeats` | For repeated k-fold cross-validation, the number of complete fold sets to compute. |
| `p` | For leave-group-out cross-validation, the training percentage. |
| `search` | Either `"grid"` or `"random"`, describing how the tuning parameter grid is determined. |
| `initialWindow`, `horizon`, `fixedWindow`, `skip` | Arguments passed to `createTimeSlices()` when `method = "timeslice"`. |
| `verboseIter` | Logical value indicating whether a training log should be printed. |
| `returnData` | Logical value indicating whether the training data should be saved. |
| `returnResamp` | Controls how much resampled summary metric output is saved. Values are `"final"`, `"all"`, or `"none"`. |
| `savePredictions` | Controls how much holdout prediction output from each resample is saved. Values are `"all"`, `"final"`, or `"none"`. The `"final"` option saves predictions for the optimal tuning parameters. |
| `classProbs` | Logical value indicating whether class probabilities should be computed for classification models. |
| `summaryFunction` | Function used to compute performance metrics across resamples. |
| `selectionFunction` | Function used to select the optimal tuning parameter. |
| `preProcOptions` | List of options passed to `preProcess()`. The preprocessing method itself is specified in the `preProc` argument of `train()`. |
| `sampling` | Optional sampling method applied after resampling, usually for class imbalance. Common values include `"none"`, `"down"`, `"up"`, `"smote"`, and `"rose"

### `tuneGrid()` Parameters
___________________________________________________________________________________________________________

The `tuneGrid` arguement utilizes the `expand.grid()` function with two nested `seq()` functions. Together, these functions work as a grid search for specific parameters. 

- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.

### Create `evaluate_classifier()`
___________________________________________________________________________________________________________

```{r}
evaluate_classifier <- function(actual, predicted, probability_yes, positive_class = "Yes") {
  actual <- factor(actual, levels = c("Yes", "No"))
  predicted <- factor(predicted, levels = c("Yes", "No"))

  cm <- caret::confusionMatrix(
    data = predicted,
    reference = actual,
    positive = positive_class
  )

  roc_obj <- pROC::roc(
    response = actual,
    predictor = probability_yes,
    levels = c("No", "Yes"),
    direction = "<",
    quiet = TRUE
  )

  precision <- cm$byClass["Precision"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall / (precision + recall)
  )

  tibble::tibble(
    accuracy = as.numeric(cm$overall["Accuracy"]),
    balanced_accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    precision = as.numeric(precision),
    sensitivity_recall = as.numeric(recall),
    specificity = as.numeric(cm$byClass["Specificity"]),
    f1_score = as.numeric(f1),
    auc = as.numeric(pROC::auc(roc_obj))
  )
}
```

### *Logistic Regression Model - Lasso Regularization*

### Determine Parameters
____________________________________________________________________________________________________________

Lasso regularization fits a logistic regression model with an L1 penalty. The L1 penalty can shrink some coefficients exactly to zero, so lasso can perform both regularization and variable selection. Because the outcome is binary, the model was fit with `family = "binomial"`. The target variable was releveled so that `Yes` is the event class used by `twoClassSummary()`.

#### Set Up `trainControl()`

```{r}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r}
lasso_grid <- expand.grid(
  alpha = 1,
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

log_lasso <- caret::train(
  loan_default ~ .,
  data = train_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = lasso_grid,
  metric = "ROC",
  family = "binomial")
```

```{r}
log_lasso
```

```{r}
log_lasso$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
lasso_prob <- predict(
  object = log_lasso,
  newdata = hold_logit,
  type = "prob"
)
```

#### Predict Class

```{r}
lasso_class <- predict(
  object = log_lasso,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
lasso_predictions <- tibble::tibble(
  actual = factor(hold_logit$loan_default, levels = c("No", "Yes")),
  predicted = factor(lasso_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(lasso_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(lasso_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
lasso_metrics <- evaluate_classifier(
  actual = lasso_predictions$actual,
  predicted = lasso_predictions$predicted,
  probability_yes = lasso_predictions$probability_yes)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(log_lasso, file.path(project_dir, "log_lasso_model.rds"))
```

#### Save Results

```{r}
lasso_results <- lasso_metrics %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Lasso",
    sample = "Full",
    alpha = log_lasso$bestTune$alpha,
    lambda = log_lasso$bestTune$lambda
  )

saveRDS(lasso_results, file.path(project_dir, "log_lasso_results.rds"))
```

## *Logistic Regression Model - Ridge Regularization*

### Determine Parameters
____________________________________________________________________________________________________________

Ridge regularization fits a logistic regression model with an L2 penalty. Unlike lasso, ridge does not usually force coefficients exactly to zero. Instead, it shrinks coefficients toward zero, which can improve stability when predictors are correlated.

#### Set Up `trainControl()`

```{r}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r}
ridge_grid <- expand.grid(
  alpha = 0,
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

log_ridge <- caret::train(
  loan_default ~ .,
  data = train_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = ridge_grid,
  metric = "ROC",
  family = "binomial"
)
```

```{r}
log_ridge
```

```{r}
log_ridge$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
ridge_prob <- predict(
  object = log_ridge,
  newdata = hold_logit,
  type = "prob"
)
```

#### Predict Class

```{r}
ridge_class <- predict(
  object = log_ridge,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
ridge_predictions <- tibble::tibble(
  actual = factor(hold_logit$loan_default, levels = c("No", "Yes")),
  predicted = factor(ridge_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(ridge_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(ridge_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r ridge}
ridge_metrics <- evaluate_classifier(
  actual = ridge_predictions$actual,
  predicted = ridge_predictions$predicted,
  probability_yes = ridge_predictions$probability_yes
)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
ridge_metrics <- evaluate_classifier(
  actual = ridge_predictions$actual,
  predicted = ridge_predictions$predicted,
  probability_yes = ridge_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(log_ridge, file.path(project_dir, "log_ridge_model.rds"))
```

#### Save Results

```{r}
ridge_results <- ridge_metrics %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Ridge",
    sample = "Full",
    alpha = log_ridge$bestTune$alpha,
    lambda = log_ridge$bestTune$lambda
  )

saveRDS(ridge_results, file.path(project_dir, "log_ridge_results.rds"))
```

## *Logistic Regression Model - Elastic Net Regularization*

### Determine Parameters
____________________________________________________________________________________________________________

Elastic net combines L1 and L2 regularization. It is useful when predictors are correlated but variable selection is still desired. The `alpha` parameter controls the balance between ridge and lasso behavior.

#### Set Up `trainControl()`

```{r}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r}
elastic_grid <- expand.grid(
  alpha = seq(0.1, 0.9, length = 5),
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

log_elastic <- caret::train(
  loan_default ~ .,
  data = train_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = elastic_grid,
  metric = "ROC",
  family = "binomial"
)
```

```{r}
log_elastic
```

```{r}
log_elastic$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
elastic_prob <- predict(
  object = log_elastic,
  newdata = hold_logit,
  type = "prob"
)
```

#### Predict Class

```{r}
elastic_class <- predict(
  object = log_elastic,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
elastic_predictions <- tibble::tibble(
  actual = factor(hold_logit$loan_default, levels = c("No", "Yes")),
  predicted = factor(elastic_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(elastic_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(elastic_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
elastic_metrics <- evaluate_classifier(
  actual = elastic_predictions$actual,
  predicted = elastic_predictions$predicted,
  probability_yes = elastic_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(log_elastic, file.path(project_dir, "log_elastic_model.rds"))
```

#### Save Results

```{r}
elastic_results <- elastic_metrics %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Elastic Net",
    sample = "Full",
    alpha = log_elastic$bestTune$alpha,
    lambda = log_elastic$bestTune$lambda
  )

saveRDS(elastic_results, file.path(project_dir, "log_elastic_results.rds"))
```

## *Logistic Regression Model - Lasso Regularization - Undersample*

### Determine Parameters
____________________________________________________________________________________________________________

#### Set Up `trainControl()`

```{r}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r}
lasso_grid_us <- expand.grid(
  alpha = 1,
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

log_lasso_us <- caret::train(
  loan_default ~ .,
  data = train_under_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = lasso_grid_us,
  metric = "ROC",
  family = "binomial")
```

```{r}
log_lasso_us
```

```{r}
log_lasso_us$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
lasso_prob_us <- predict(
  object = log_lasso_us,
  newdata = hold_logit,
  type = "prob"
)
```

#### Predict Class

```{r}
lasso_class_us <- predict(
  object = log_lasso_us,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
lasso_predictions_us <- tibble::tibble(
  actual = factor(hold_logit$loan_default, levels = c("No", "Yes")),
  predicted = factor(lasso_class_us, levels = c("No", "Yes")),
  probability_yes = as.numeric(lasso_prob_us[, "Yes"])
)
```

#### Preview Table

```{r}
head(lasso_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
lasso_metrics_us <- evaluate_classifier(
  actual = lasso_predictions_us$actual,
  predicted = lasso_predictions_us$predicted,
  probability_yes = lasso_predictions_us$probability_yes)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(log_lasso_us, file.path(project_dir, "log_lasso_undersampled_model.rds"))
```

#### Save Results

```{r}
lasso_results_us <- lasso_metrics_us %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Lasso",
    sample = "Undersampled",
    alpha = log_lasso$bestTune$alpha,
    lambda = log_lasso$bestTune$lambda
  )

saveRDS(lasso_results_us, file.path(project_dir, "log_lasso_undersampled_results.rds"))
```

## *Logistic Regression Model - Ridge Regularization - Undersample*

### Determine Parameters
____________________________________________________________________________________________________________

Ridge regularization fits a logistic regression model with an L2 penalty. Unlike lasso, ridge does not usually force coefficients exactly to zero. Instead, it shrinks coefficients toward zero, which can improve stability when predictors are correlated.

#### Set Up `trainControl()`

```{r}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r}
ridge_grid_us <- expand.grid(
  alpha = 0,
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

log_ridge_us <- caret::train(
  loan_default ~ .,
  data = train_under_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = ridge_grid_us,
  metric = "ROC",
  family = "binomial"
)
```

```{r}
log_ridge_us
```

```{r}
log_ridge_us$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
ridge_prob_us <- predict(
  object = log_ridge_us,
  newdata = hold_logit,
  type = "prob"
)
```

#### Predict Class

```{r}
ridge_class_us <- predict(
  object = log_ridge_us,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
ridge_predictions_us <- tibble::tibble(
  actual = factor(hold_logit$loan_default, levels = c("No", "Yes")),
  predicted = factor(ridge_class_us, levels = c("No", "Yes")),
  probability_yes = as.numeric(ridge_prob_us[, "Yes"]))
```

#### Preview Table

```{r}
head(ridge_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
ridge_metrics_us <- evaluate_classifier(
  actual = ridge_predictions_us$actual,
  predicted = ridge_predictions_us$predicted,
  probability_yes = ridge_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(log_ridge_us, file.path(project_dir, "log_ridge_undersampled_model.rds"))
```

#### Save Results

```{r}
ridge_results_us <- ridge_metrics_us %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Ridge",
    sample = "Undersampled",
    alpha = log_ridge_us$bestTune$alpha,
    lambda = log_ridge_us$bestTune$lambda
  )

saveRDS(ridge_results_us, file.path(project_dir, "log_ridge_undersampled_results.rds"))
```

## *Logistic Regression Model - Elastic Net Regularization - Undersample*

#### Set Up `trainControl()`

```{r elastic}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r elastic}
elastic_grid_us <- expand.grid(
  alpha = seq(0.1, 0.9, length = 5),
  lambda = seq(0.001, 0.1, length = 5)
)
```

### Fit Model
____________________________________________________________________________________________________________

```{r elastic}
set.seed(123)

log_elastic_us <- caret::train(
  loan_default ~ .,
  data = train_under_logit,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = elastic_grid_us,
  metric = "ROC",
  family = "binomial"
)
```

```{r elastic}
log_elastic_us
```

```{r elastic}
log_elastic_us$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________


#### Predict Probabilities

```{r elastic}
elastic_prob_us <- predict(
  model = log_elastic_us,
  newdata = hold_logit,
  type = "prob"
)[, "Yes"]
```

#### Predict Class

```{r elastic}
elastic_pred_us <- predict(
  model = log_elastic_us,
  newdata = hold_logit,
  type = "raw"
)
```

#### Tabulate Predictions

```{r elastic}
elastic_predictions_us <- tibble::tibble(
  actual = hold_logit$loan_default,
  predicted = elastic_pred_us,
  probability_yes = elastic_prob_us
)
```

#### Preview Table

```{r elastic}
head(elastic_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r elastic}
elastic_metrics_us <- evaluate_classifier(
  actual = elastic_predictions_us$actual,
  predicted = elastic_predictions_us$predicted,
  probability_yes = elastic_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r elastic}
saveRDS(log_elastic_us, file.path(project_dir, "log_elastic_undersampled_model.rds"))
```

#### Save Results

```{r ridge}
elastic_results_us <- elastic_metrics_us %>%
  mutate(
    model_type = "Logistic Regression",
    regularization = "Elastic Net",
    sample = "Undersampled",
    alpha = log_elastic_us$bestTune$alpha,
    lambda = log_elastic_us$bestTune$lambda
  )

saveRDS(elastic_results_us file.path(project_dir, "log_elastic_undersampled_results.rds"))
```