---
title: "Binary Classification of Loan Default Using Machine Learning"
author: "storeraa17"
data: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
    The abstract should be made of
    - purpose
    - main methods
    - experimental results
    - key discussion points
    - final conclusions
keywords:
    - Predictive Statistics
    - Binary Classification
    - Machine Learning
    - Model Evaluation
    - Data Cleaning
    - Data Processing
    - Data Visualization
output:
  html_document:
    toc: true
    toc_depth: 7
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
    fig_width: 8
    fig_height: 5
    fig_caption: true
    self_contained: true
  word_document:
    toc: true
    fig_caption: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

---

# **Introduction**

The purpose of this project is to support better loan risk assessment. A reliable default prediction model can help lenders make more informed decisions, reduce financial losses, and improve consistency in loan evaluation. Rather than relying only on manual review or a limited set of business rules, a predictive model can evaluate many borrower and loan characteristics at the same time. Thus, the baility to predict loan default is a critical problem. For example, financial analysts may need to estimate the risk that a borrow will fail to repay a loan. When a loan is issued, the lender accepts financial risk based on information about the borrower, the loan terms, credit history, income, debt obligations, and application characteristics. If risk is underestimated, the lender may approve loans that are more likely to default. If risk is overestimated, the lender may reject borrowers who could have successfully repaid their loans.  Herein, we report a machine learning model used for predictive binary classification of loan default.

The machine learning model used to evaluated will be subjected to several inputs to supply the desired loan default classification output. The make-up inputs cover several important features to consider, such as borrow information, loan terms, finances, credit history, accounting, delinquencies, administration, and status variables. The data representing these variables is split into two datasets:

1. `train.csv`
2. 'holdout.csv`

The dataset 'train.csv` represents the training data and will be the dataset used for model learning. While, `holdout.csv` will be the dataset that the model is evaluated on. The evaluation of the `holdout.csv` dataset will be done on the premise of F1 score primarily. However, F1 score is not exclusively the only metric used to determine model performance. Secondary evaluation metrics used to examine model performance include accuracy, balanced accuracy, recall/sensitivity, precision, number of false positives, number of false negatives, and AUC of an ROC curve.

These performance metrics will be used to evaluate decision trees, tree ensembles, logistic regression, and neural network models. The constructions of these models can be for in the `model` folder. Nevertheless, the best resulting models, based on F1 score, will be imported and model performances will be compared to determine the optimal model.

The variables found in `train.csv` and `holdout.csv` are described in the data dictionary below:

| Variable | Type | Description |
|:---|:---|:---|
| `loan_default` | Categorical | Indicates whether the borrower defaulted. Values are `Yes` and `No`. |
| `emp_title` | Categorical / text | Borrower's reported job title. High-cardinality text variable. |
| `emp_length` | Categorical / ordinal | Borrower's reported employment length. |
| `home_ownership` | Categorical | Borrower's home ownership status. |
| `annual_inc` | Numeric | Borrower's self-reported annual income. |
| `verification_status` | Categorical | Whether income or financial information was verified. |
| `address` | Categorical / text | Borrower's recorded address. High-cardinality text variable. |
| `loan_amnt` | Numeric | Original loan principal amount. |
| `term` | Categorical | Scheduled repayment length, generally 36 or 60 months. |
| `int_rate` | Numeric | Annual interest rate assigned to the loan. |
| `installment` | Numeric | Scheduled monthly payment. |
| `grade` | Categorical / ordinal | Broad lender-assigned credit grade. |
| `sub_grade` | Categorical / ordinal | More detailed credit grade. |
| `purpose` | Categorical | Borrower's stated loan purpose. |
| `title` | Categorical / text | Borrower-provided title for the loan request. |
| `dti` | Numeric | Debt-to-income ratio. |
| `earliest_cr_line` | Date-like character | Date of borrower's earliest reported credit line. |
| `fico_range_low` | Numeric | Lower bound of reported FICO range. |
| `fico_range_high` | Numeric | Upper bound of reported FICO range. |
| `inq_last_6mths` | Numeric | Number of credit inquiries in the last six months. |
| `mths_since_last_delinq` | Numeric | Months since most recent delinquency. Missing values may indicate no prior delinquency or unavailable information. |
| `last_credit_pull_d` | Date-like character | Most recent credit pull date. |
| `open_acc` | Numeric | Number of open credit accounts. |
| `pub_rec` | Numeric | Number of derogatory public records. |
| `revol_bal` | Numeric | Total revolving balance. |
| `revol_util` | Numeric | Revolving credit utilization rate. |
| `total_acc` | Numeric | Total number of credit accounts. |
| `mort_acc` | Numeric | Number of mortgage accounts. |
| `pub_rec_bankruptcies` | Numeric | Number of public-record bankruptcies. |
| `delinq_2yrs` | Numeric | Number of delinquency incidents in the last two years. |
| `acc_now_delinq` | Numeric | Number of accounts currently delinquent. |
| `issue_d` | Date-like character | Month and year the loan was issued. |
| `initial_list_status` | Categorical | Initial listing status assigned to the loan. |
| `application_type` | Categorical | Individual, joint, or direct-pay application type. |
| `hardship_flag` | Categorical | Indicates hardship arrangement status. |
| `debt_settlement_flag` | Categorical | Indicates debt-settlement status. |

---

# **Methods**

## *Data Loading*

The datasets, `train.csv` and `holdout.csv`, were loaded and the target variable. `loan_default`, was immediately converted into a factor.

```{r data loading}
train_raw <- readr::read_csv(
  file.path(project_dir, "train.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

```{r data loading}
hold_raw <- readr::read_csv(
  file.path(project_dir, "holdout.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

## *Data Inspection*

### Previewing the Dataset
____________________________________________________________________________________________________________

The structure of the datasets were assessed using the `str()` function. Subsequent previewing of the datasets was done through the utilization of the `head()` function, which shows the first few rows of a dataset.

```{r data inspection}
str(train_raw)
```

```{r data inspection}
str(hold_raw)
```

```{r data inspection}
head(train_raw)
```

```{r data inspection}
head(hold_raw)
```

### Dataset Dimensions
____________________________________________________________________________________________________________

The dimensions of the dataset were determined by employed the `dim()` function.

```{r data inspection}
dim(train_raw)
```

```{r data inspection}
dim(hold_raw)
```

### The Number of Unique Instances
____________________________________________________________________________________________________________
The number of unique instances for each variable were examined via `sapply()` and `n_distinct()` functions.

```{r data inspection}
sapply(train_raw, n_distinct)
```

```{r data inspection}
sapply(hold_raw, n_distinct)
```

### Class Comparison of Variables
____________________________________________________________________________________________________________
It is important to compare the classes of each variable in the training and holdout datasets to ensure consistency. This can be done using the `sapply()` and `class()` functions. To present the findings more cleanly, a tibble can be created to display the variable names, their corresponding classes in the training dataset, and their classes in the holdout dataset. The function `tibble::tibble()` can be used to create this tibble, and the `sapply()` function can be used to apply the `class()` function to each variable in both datasets. The resulting tibble will provide a clear comparison of the variable classes between the two datasets.

```{r data inspection}
class_comparison <- tibble::tibble(
  Variable = names(train_raw),
  Train_Class = sapply(train_raw, function(x) class(x)[1]),
  Holdout_Class = sapply(hold_raw, function(x) class(x)[1])
)

class_comparison
```

## *Summary Statistics*

### Summary Statistics
____________________________________________________________________________________________________________

The summary statistics of the datasets were examined using the `skimr::skim()` function. This function provides a comprehensive overview of the data, including the number of missing values, mean, median, standard deviation, and other relevant statistics for each variable.

```{r data }
skimr::skim(train_raw)
```

```{r data processing}
skimr::skim(hold_raw)
```

## *Data Visualization**

### Grouping Numeric Variables
____________________________________________________________________________________________________________

The numeric variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data visualization}
numeric_vars <- train_raw %>%
  select(where(is.numeric)) %>%
  names()

numeric_vars
```

The numeric variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data visualization}
hold_numeric_vars <- hold_raw %>%
  select(where(is.numeric)) %>%
  names()

hold_numeric_vars
```

### Grouping Categorical Variables
____________________________________________________________________________________________________________

The categorical variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data visualization}
categorical_vars <- train_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

categorical_vars
```

The categorical variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data visualization}
hold_categorical_vars <- hold_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

hold_categorical_vars
```

### Target Variable Distribution
____________________________________________________________________________________________________________

#### Distribution of the Target Variable in the Training Dataset

The distribution of the target variable, `loan_default`, in the training dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data visualization}
train_target_distribution <- table(train_raw$loan_default)
train_target_proportions <- prop.table(train_target_distribution)
```

```{r data visualization}
train_target_distribution_df <- as.data.frame(train_target_distribution)
colnames(train_target_distribution_df) <- c("loan_default", "Count")
```

```{r data visualization}
ggplot(train_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Training Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
```

#### Distribution of the Target Variable in the Holdout Dataset

The distribution of the target variable, `loan_default`, in the holdout dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data visualization}
hold_target_distribution <- table(hold_raw$loan_default)
hold_target_proportions <- prop.table(hold_target_distribution)
```

```{r data visualization}
hold_target_distribution_df <- as.data.frame(hold_target_distribution)
colnames(hold_target_distribution_df) <- c("loan_default", "Count")
```

```{r data visualization}
ggplot(hold_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity", color="white", alpha=0.9) +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Holdout Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
```

### Numeric Variable Distributions - Histograms
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using histograms to visualize their distributions. The `DataExplorer` package was used to create the histograms, with customizations applied to the theme and appearance of the plots.

#### Historgrams of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_histogram(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Historgrams of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_histogram(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Boxplots
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using boxplots to visualize their distributions and identify potential outliers. The `DataExplorer` package was used to create the boxplots, with customizations applied to the theme and appearance of the plots.

#### Boxplots of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_boxplot(
  train_raw %>%
    select(loan_default, all_of(numeric_vars)),
  by = "loan_default",
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_boxplot_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Boxplots of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_boxplot(
  hold_raw %>%
    select(loan_default, all_of(hold_numeric_vars)),
  by = "loan_default",
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_boxplot_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Density Plots
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using density plots to visualize their distributions. The `DataExplorer` package was used to create the density plots, with customizations applied to the theme and appearance of the plots.

#### Density Plots of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_density(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_density_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Density Plots of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_density(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_density_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Heat Map
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using correlation heat maps to visualize the relationships between the variables. The `DataExplorer` package was used to create the heat maps, with customizations applied to the theme and appearance of the plots.

#### Correlation Heat Map of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_correlation(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  cor_args = list(
    use = "pairwise.complete.obs",
    method = "pearson"
  )
)
```

#### Correlation Heat Map of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_correlation(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  cor_args = list(
    use = "pairwise.complete.obs",
    method = "pearson"
  )
)
```

### Numeric Variable Distributions - Correlation Pairs
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using correlation matrices to visualize the relationships between the variables. The `cor()` function was used to calculate the correlation coefficients, with the `use` argument set to "pairwise.complete.obs" to handle missing values. The resulting correlation matrices were then visualized using heat maps created with the `corrplot` package.

#### Correlation Matrix of Numeric Variables in the Training Dataset

```{r data visualization}
correlation_matrix_train <- cor(train_raw %>% select(all_of(numeric_vars)), use = "pairwise.complete.obs")
```

##### Correlation Matrix of Numeric Variables in the Holdout Dataset
```{r data visualization}
correlation_matrix_hold <- cor(hold_raw %>% select(all_of(hold_numeric_vars)), use = "pairwise.complete.obs")
```

### Categorical Variable Distributions - Frequency Tables
____________________________________________________________________________________________________________

The categorical variables in the training and holdout datasets were examined using frequency tables to visualize their distributions. The `table()` function was used to create the frequency tables, with customizations applied to the theme and appearance of the tables.

#### Frequecy Tables of Categorical Variables in the Training Dataset

```{r data visualization}
for (col in categorical_vars) {
  print(paste("Top categories for", col, "in Training Dataset:"))
  print(head(sort(table(train_raw[[col]]), decreasing = TRUE), 5))
}
```

#### Frequecy Tables of Categorical Variables in the Holdout Dataset

```{r data visualization}
for (col in hold_categorical_vars) {
  print(paste("Top categories for", col, "in Holdout Dataset:"))
  print(head(sort(table(hold_raw[[col]]), decreasing = TRUE), 5))
}
```

### Categorical Variable Distributions - Bar Plots
____________________________________________________________________________________________________________

The categorical variables in the training and holdout datasets were examined using bar plots to visualize their distributions. The `ggplot2` package was used to create the bar plots, with customizations applied to the theme and appearance of the plots. A function was created to generate bar plots for each categorical variable in the datasets. The function takes three arguments: the dataset, the list of categorical variables, and a title prefix for the plots. It iterates through each categorical variable, creates a bar plot using `ggplot2`, and prints the plot.

#### Creating the Bar Plots Function

```{r data visualization}
bar_plots <- function(data, categorical_vars, title_prefix, max_levels = 20) {
  for (col in categorical_vars) {

    n_levels <- data %>%
      distinct(.data[[col]]) %>%
      nrow()

    if (n_levels > max_levels) {
      message("Skipping ", col, ": ", n_levels, " unique levels.")
      next
    }

    plot <- ggplot(data, aes(x = .data[[col]])) +
      geom_bar(fill = "#4E79A7", color = "white", alpha = 0.9) +
      labs(
        title = paste(title_prefix, col),
        x = col,
        y = "Count"
      ) +
      theme_bw(base_size = 14) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )

    print(plot)
  }
}
```

#### Bar Plots of Categorical Variables in the Training Dataset

```{r data visualization}
bar_plots(
  train_raw,
  categorical_vars,
  "Distribution of Categorical Variable in Training Dataset:"
)
```

#### Bar Plots of Categorical Variables in the Holdout Dataset

```{r data visualization}
bar_plots(
  hold_raw,
  hold_categorical_vars,
  "Distribution of Categorical Variable in Holdout Dataset:"
)
```

### Categorial Variable Distributions - Proportions by Loan Default Status
____________________________________________________________________________________________________________

#### Proportions of Categorical Variables by Loan Default Status in the Training Dataset

```{r data visualization}
for (col in categorical_vars) {
  prop_table <- prop.table(
    table(train_raw[[col]], train_raw$loan_default),
    margin = 1
  )

  print(paste("Proportions of", col, "by Loan Default Status in Training Dataset:"))
  print(prop_table)
}
```

#### Proportions of Categorical Variables by Loan Default Status in the Holdout Dataset

```{r data visualization}
for (col in hold_categorical_vars) {
  prop_table <- prop.table(
    table(hold_raw[[col]], hold_raw$loan_default),
    margin = 1
  )

  print(paste("Proportions of", col, "by Loan Default Status in Holdout Dataset:"))
  print(prop_table)
}
```

## *Data Quality Assessment*

### Missing Value Counts
____________________________________________________________________________________________________________

The presence of missing values in the training and holdout datasets was examined using the `sapply()` function to apply the `n_missing()` function to each variable. The `n_missing()` function counts the number of missing values in a variable. The results were then presented in a table format for easy interpretation.

#### Missing Values in the Training Dataset

```{r dqa}
sapply(train_raw, n_missing)
```

#### Missing Values in the Holdout Dataset

```{r dqa}
sapply(hold_raw, n_missing)
```

### Missing Value Plots
____________________________________________________________________________________________________________

The missing values in the training and holdout datasets were visualized using the `plot_missing()` function from the `DataExplorer` package. This function creates a plot that shows the proportion of missing values for each variable in the dataset. The plots were customized using the `ggtheme` argument to apply a specific theme and appearance. The percentage of missing values for each variable is displayed on the y-axis, while the variable names are displayed on the x-axis. The plots provide a visual representation of the extent of missing data in the datasets, which can inform decisions regarding data cleaning and preprocessing.

#### Missing Value Plots in the Training Dataset

```{r dqa}
DataExplorer::plot_missing(train_raw, ggtheme = theme_bw(base_size = 14))
```

#### Missing Value Plots in the Holdout Dataset

```{r dqa}
DataExplorer::plot_missing(hold_raw, ggtheme = theme_bw(base_size = 14))
```

### Anomalic Values Assessment
____________________________________________________________________________________________________________

The strategy employed to detect anomalic values in the numeric variables of the training and holdout datasets was based on the interquartile range (IQR) method. The IQR is a measure of statistical dispersion, which is the difference between the third quartile (Q3) and the first quartile (Q1). Anomalic values, or outliers, were identified as any data points that fall below Q1 - 1.5 * IQR or above Q3 + 1.5 * IQR. The `boxplot.stats()` function was used to calculate the IQR and identify outliers for each numeric variable in both datasets. The results were then printed for each variable that contained anomalic values. This only applies to numeric variables, as categorical variables do not have a meaningful IQR. Outliers are expected in financial data. High income, high revolving balance, and large loan amounts can be valid borrower records rather than data entry errors. For this reason, the workflow avoids deleting outliers by default. The model search includes variants that add outlier flags and log transformations.

#### Anomalic Values in the Training Dataset

```{r dqa}
for (col in numeric_vars) {
  outliers <- boxplot.stats(train_raw[[col]])$out

  if (length(outliers) > 0) {
    print(paste("Anomalous values in", col, "in Training Dataset:"))
    print(head(outliers, 20))
  }
}
```

#### Anomalic Values in the Holdout Dataset

```{r dqa}
for (col in hold_numeric_vars) {
  outliers <- boxplot.stats(hold_raw[[col]])$out

  if (length(outliers) > 0) {
    print(paste("Anomalous values in", col, "in Holdout Dataset:"))
    print(head(outliers, 20))
  }
}
```

### Inconsistent Values Assessment
_____________________________________________________________________________________________________________

First, the categorical variables in both the training and holdout datasets were checked for inconsistent values. Inconsistent values refer to unexpected or invalid entries in categorical variables that do not conform to the expected categories. The `unique()` function was used to identify the unique values present in each categorical variable. If a variable had more than 10 unique values, it was flagged for further inspection, as this may indicate potential inconsistencies or data quality issues. The results were printed for each variable that contained inconsistent values. Second, the numeric variables in both datasets were checked for inconsistent values. Inconsistent values in numeric variables refer to unexpected or invalid entries that fall outside the expected range or distribution of values. The `summary()` function was used to examine the summary statistics of each numeric variable, including the minimum, maximum, mean, and quartiles. If a variable had extreme values or outliers that were significantly different from the rest of the data, it was flagged for further inspection. The results were printed for each variable that contained inconsistent values. This step is important to ensure the integrity and reliability of the data before proceeding with further analysis and modeling.

#### Inconsistent Values in Categorical Variables in the Training Dataset

```{r dqa}
for (col in categorical_vars) {
  unique_values <- unique(train_raw[[col]])

  if (length(unique_values) > 10) {
    print(paste("High-cardinality or potentially inconsistent values in", col, "in Training Dataset:"))
    print(head(unique_values, 20))
  }
}
```

#### Inconsistent Values in Categorical Variables in the Holdout Dataset

```{r dqa}
for (col in hold_categorical_vars) {
  unique_values <- unique(hold_raw[[col]])

  if (length(unique_values) > 10) {
    print(paste("High-cardinality or potentially inconsistent values in", col, "in Holdout Dataset:"))
    print(head(unique_values, 20))
  }
}
```

#### Inconsistent Values in Numeric Variables in the Training Dataset

```{r dqa}
for (col in numeric_vars) {
  summary_stats <- summary(train_raw[[col]])

  if (summary_stats["Min."] < 0 || summary_stats["Max."] > 1000000) {
    print(paste("Potentially inconsistent numeric values in", col, "in Training Dataset:"))
    print(summary_stats)
  }
}
```

#### Inconsistent Values in Numeric Variables in the Holdout Dataset

```{r dqa}
for (col in hold_numeric_vars) {
  summary_stats <- summary(hold_raw[[col]])

  if (summary_stats["Min."] < 0 || summary_stats["Max."] > 1000000) {
    print(paste("Potentially inconsistent numeric values in", col, "in Holdout Dataset:"))
    print(summary_stats)
  }
}
```

### High Cardinality Assessment
_____________________________________________________________________________________________________________

A categorical variable with a high cardinality is one that has a large number of unique categories or levels. High cardinality can pose challenges for modeling, as it may lead to overfitting or increased computational complexity. In this analysis, categorical variables with more than 10 unique values were flagged as high cardinality. The `unique()` function was used to identify the unique values present in each categorical variable, and the results were printed for each variable that had high cardinality. These variables can be further examined to determine if they can be transformed or encoded in a way that reduces their cardinality, such as through grouping or feature engineering. The presence of high cardinality categorical variables during modeling can be addressed through techniques such as one-hot encoding, target encoding, or embedding representations, depending on the modeling approach and the specific characteristics of the data. The high cardinality variables can limit the choice of models and may require additional preprocessing steps to ensure that the model can effectively learn from the data.

#### High Cardinality Categorical Variables in the Training Dataset

```{r dqa}
for (col in categorical_vars) {
  unique_values <- unique(train_raw[[col]])

  if (length(unique_values) > 10) {
    print(paste("High cardinality categorical variable in Training Dataset:", col))
    print(head(unique_values, 20))
  }
}
```

#### High Cardinality Categorical Variables in the Holdout Dataset

```{r dqa}
for (col in hold_categorical_vars) {
  unique_values <- unique(hold_raw[[col]])

  if (length(unique_values) > 10) {
    print(paste("High cardinality categorical variable in Holdout Dataset:", col))
    print(head(unique_values, 20))
  }
}
```

### Rare Category Assessment
_____________________________________________________________________________________________________________

The presence of rare categories can be assessed with code below. A function, `assess_rare_categories`, was created. This function has five dependencies in order to properly evaluate the presence of rare categories.

#### Creating A Rare Category Assessment Function

```{r dqa}
assess_rare_categories <- function(data,
                                   variables = NULL,
                                   target = "loan_default",
                                   min_count = 30,
                                   min_prop = 0.01) {
  if (is.null(variables)) {
    variables <- names(data)[sapply(data, function(x) is.character(x) || is.factor(x))]
  }

  variables <- setdiff(variables, target)

  purrr::map_dfr(variables, function(v) {
    data %>%
      transmute(
        variable = v,
        value = as.character(.data[[v]])
      ) %>%
      mutate(
        value = ifelse(is.na(value) | stringr::str_squish(value) == "", "Missing", value),
        value = stringr::str_squish(value)
      ) %>%
      count(variable, value, name = "count") %>%
      mutate(
        total_rows = nrow(data),
        proportion = count / total_rows,
        rare_category = count < min_count | proportion < min_prop
      )
  })
}
```

#### Rare Category Assessment of Categorical Variables in the Training Dataset

```{r dqa}
train_rare_categories <- assess_rare_categories(
  data = train_raw,
  target = "loan_default",
  min_count = 30,
  min_prop = 0.01
)

train_rare_summary <- train_rare_categories %>%
  group_by(variable) %>%
  summarise(
    total_levels = n(),
    rare_levels = sum(rare_category),
    percent_levels_rare = rare_levels / total_levels,
    rows_in_rare_levels = sum(count[rare_category]),
    percent_rows_in_rare_levels = rows_in_rare_levels / first(total_rows),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_rows_in_rare_levels), desc(rare_levels))

train_rare_summary %>%
  mutate(
    percent_levels_rare = scales::percent(percent_levels_rare, accuracy = 0.1),
    percent_rows_in_rare_levels = scales::percent(percent_rows_in_rare_levels, accuracy = 0.1)
  ) %>%
  knitr::kable(caption = "Rare Category Assessment for Training Data")
```

#### Rare Category Assessment of Categorical Variables in the Holdout Dataset

```{r dqa}
holdout_rare_categories <- assess_rare_categories(
  data = hold_raw,
  target = "loan_default",
  min_count = 30,
  min_prop = 0.01
)

holdout_rare_summary <- holdout_rare_categories %>%
  group_by(variable) %>%
  summarise(
    total_levels = n(),
    rare_levels = sum(rare_category),
    percent_levels_rare = rare_levels / total_levels,
    rows_in_rare_levels = sum(count[rare_category]),
    percent_rows_in_rare_levels = rows_in_rare_levels / first(total_rows),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_rows_in_rare_levels), desc(rare_levels))

holdout_rare_summary %>%
  mutate(
    percent_levels_rare = scales::percent(percent_levels_rare, accuracy = 0.1),
    percent_rows_in_rare_levels = scales::percent(percent_rows_in_rare_levels, accuracy = 0.1)
  ) %>%
  knitr::kable(caption = "Rare Category Assessment for Holdout Data")
```

### Data Leakage Assessment
_____________________________________________________________________________________________________________

It was assumed that data leakage was not a potential problem. However, this will not be the case for real-world examples. Both the training and holdout datasets were examined for possible data leakages.

#### Create Data Leakage Assessment Function

```{r dqa}
assess_data_leakage <- function(data,
                                target = "loan_default",
                                positive_class = "Yes",
                                negative_class = "No") {
  leakage_name_patterns <- c(
    "settlement", "hardship", "recover", "collection", "last",
    "payment", "pymnt", "status", "default", "chargeoff",
    "out_prncp", "total_rec", "funded"
  )

  predictors <- setdiff(names(data), target)

  name_flags <- tibble::tibble(
    variable = predictors,
    leakage_name_flag = stringr::str_detect(
      stringr::str_to_lower(variable),
      paste(leakage_name_patterns, collapse = "|")
    )
  )

  categorical_results <- purrr::map_dfr(predictors, function(v) {
    if (!(is.character(data[[v]]) || is.factor(data[[v]]))) return(NULL)

    temp <- data %>%
      mutate(
        target_value = .data[[target]] == positive_class,
        predictor_value = as.character(.data[[v]]),
        predictor_value = ifelse(is.na(predictor_value), "Missing", predictor_value)
      ) %>%
      group_by(predictor_value) %>%
      summarise(
        n = n(),
        default_rate = mean(target_value, na.rm = TRUE),
        .groups = "drop"
      )

    tibble::tibble(
      variable = v,
      variable_type = "categorical",
      max_default_rate = max(temp$default_rate, na.rm = TRUE),
      min_default_rate = min(temp$default_rate, na.rm = TRUE),
      default_rate_spread = max_default_rate - min_default_rate,
      levels = nrow(temp),
      suspicious_strength_flag = default_rate_spread >= 0.75
    )
  })

  numeric_results <- purrr::map_dfr(predictors, function(v) {
    if (!is.numeric(data[[v]])) return(NULL)

    temp <- data %>%
      select(all_of(c(target, v))) %>%
      filter(!is.na(.data[[target]]), !is.na(.data[[v]]))

    if (nrow(temp) == 0 || length(unique(temp[[v]])) < 2) return(NULL)

    roc_obj <- tryCatch(
      pROC::roc(
        response = temp[[target]],
        predictor = temp[[v]],
        levels = c(negative_class, positive_class),
        direction = "<",
        quiet = TRUE
      ),
      error = function(e) NULL
    )

    if (is.null(roc_obj)) return(NULL)

    auc_value <- as.numeric(pROC::auc(roc_obj))
    auc_value <- max(auc_value, 1 - auc_value)

    tibble::tibble(
      variable = v,
      variable_type = "numeric",
      auc = auc_value,
      suspicious_strength_flag = auc_value >= 0.90
    )
  })

  bind_rows(categorical_results, numeric_results) %>%
    left_join(name_flags, by = "variable") %>%
    mutate(
      possible_leakage_flag = leakage_name_flag | suspicious_strength_flag
    ) %>%
    arrange(desc(possible_leakage_flag), desc(suspicious_strength_flag))
}
```

#### Data Laakage Assessment of Training Dataset

```{r dqa}
train_leakage_assessment <- assess_data_leakage(
  data = train_raw,
  target = "loan_default",
  positive_class = "Yes",
  negative_class = "No"
)

train_leakage_assessment %>%
  filter(possible_leakage_flag) %>%
  mutate(
    default_rate_spread = ifelse(
      is.na(default_rate_spread),
      NA,
      scales::percent(default_rate_spread, accuracy = 0.1)
    ),
    auc = round(auc, 3)
  ) %>%
  knitr::kable(caption = "Potential Data Leakage Flags in Training Data")
```

#### Data Leakage Assessment of Holdout Dataset

```{r dqa}
holdout_leakage_assessment <- assess_data_leakage(
  data = hold_raw,
  target = "loan_default",
  positive_class = "Yes",
  negative_class = "No"
)

holdout_leakage_assessment %>%
  filter(possible_leakage_flag) %>%
  mutate(
    default_rate_spread = ifelse(
      is.na(default_rate_spread),
      NA,
      scales::percent(default_rate_spread, accuracy = 0.1)
    ),
    auc = round(auc, 3)
  ) %>%
  knitr::kable(caption = "Potential Data Leakage Flags in Holdout Data")
```

## *Data Preprocessing*

### Character-to-Factor Conversion
_____________________________________________________________________________________________________________

Categorical fields such as `emp_length`, `home_ownership`, `verification_status`, `purpose`, `initial_list_status`, and `application_type` are treated as categorical predictors. Some categorical fields are also converted into ordered or numeric scores when their ordering has business meaning.

#### Employee Length

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    emp_length = forcats::fct_na_value_to_level(factor(emp_length), level = "Missing"),
    emp_length_numeric = case_when(
        emp_length == "< 1 year" ~ 0,
        emp_length == "1 year" ~ 1,
        emp_length == "2 years" ~ 2,
        emp_length == "3 years" ~ 3,
        emp_length == "4 years" ~ 4,
        emp_length == "5 years" ~ 5,
        emp_length == "6 years" ~ 6,
        emp_length == "7 years" ~ 7,
        emp_length == "8 years" ~ 8,
        emp_length == "9 years" ~ 9,
        emp_length == "10+ years" ~ 10,
        TRUE ~ NA_real_
      )
  )
```

#### Home Ownership

```{r data preprocessing}
train_raw$home_ownership <- as.factor(train_raw$home_ownership)
```

#### Verification Status

```{r data preprocessing}
train_raw$verification_status <- as.factor(train_raw$verification_status)
```

#### Purpose

```{r data preprocessing}
train_raw$purpose <- as.factor(train_raw$purpose)
```

#### Initial List Status

```{r data preprocessing}
train_raw$initial_list_status <- as.factor(train_raw$initial_list_status)
```

#### Application Type

```{r data preprocessing}
train_raw$application_type <- as.factor(train_raw$application_type)
```

#### Hardship Flag

```{r data preprocessing}
train_raw$hardship_flag <- as.factor(train_raw$hardship_flag)
```

#### Debt Settlement Flag

```{r data preprocessing}
train_raw$debt_settlement_flag <- as.factor(train_raw$debt_settlement_flag)
```


### Character-to-Numeric Conversion
_____________________________________________________________________________________________________________

#### Term

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    term = readr::parse_number(as.character(term))
  )
```

### Character-to-Date Conversion
_____________________________________________________________________________________________________________

The variables `issue_d`,  `earliest_cr_line`, and `last_credit_pull_d` were parsed using `lubridate`::`my()`. These parsed dates support credit history and loan timing features.

#### Issue Date

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    issue_d = lubridate::my(issue_d)
  )
```

#### Earliest Credit Line

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    earliest_cr_line = lubridate::my(earliest_cr_line)
  )
```

#### Last Credit Pull Date

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    last_credit_pull_d = lubridate::my(last_credit_pull_d)
  )
```


### Missing Value Treatment
_____________________________________________________________________________________________________________

Median imputation was employed as the imputation strategy to handle missing values. Later, this will be done on the holdout dataset. The training data median will be used on the holdout dataset as is standard protocol.

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    no_delinq_history_flag = ifelse(is.na(mths_since_last_delinq), 1, 0),
    mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 999, mths_since_last_delinq))

num_cols_train <- names(train_raw)[sapply(train_raw, is.numeric)]

train_medians <- sapply(train_raw[num_cols_train], median, na.rm = TRUE)

for (v in num_cols_train) {
  train_raw[[v]][!is.finite(train_raw[[v]])] <- NA_real_
  train_raw[[v]][is.na(train_raw[[v]])] <- train_medians[[v]]
}
```

### Handling Rare or Sparse Categories
_____________________________________________________________________________________________________________

#### Application Type

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    application_type = forcats::fct_collapse(
      factor(application_type),
      OTHER = c("JOINT", "DIRECT_PAY")
    )
  )
```

#### Home Ownership

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    home_ownership = forcats::fct_collapse(
      factor(home_ownership),
      OTHER = c("OTHER", "ANY", "NONE")
    )
  )
```

### Removing Unnecessary Variables
_____________________________________________________________________________________________________________

```{r data preprocessing}
train_raw <- train_raw %>%
  select(
    -emp_title,
    -title
  )
```


### Applying Preprocessing to Holdout Data
_____________________________________________________________________________________________________________

#### Character-to-Factor Conversion


##### Employee Length

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    emp_length = forcats::fct_na_value_to_level(factor(emp_length), level = "Missing"),
    emp_length_numeric = case_when(
        emp_length == "< 1 year" ~ 0,
        emp_length == "1 year" ~ 1,
        emp_length == "2 years" ~ 2,
        emp_length == "3 years" ~ 3,
        emp_length == "4 years" ~ 4,
        emp_length == "5 years" ~ 5,
        emp_length == "6 years" ~ 6,
        emp_length == "7 years" ~ 7,
        emp_length == "8 years" ~ 8,
        emp_length == "9 years" ~ 9,
        emp_length == "10+ years" ~ 10,
        TRUE ~ NA_real_
      )
  )
```

##### Home Ownership

```{r data preprocessing}
hold_raw$home_ownership <- as.factor(hold_raw$home_ownership)
```

##### Verification Status

```{r data preprocessing}
hold_raw$verification_status <- as.factor(hold_raw$verification_status)
```

##### Purpose

```{r data preprocessing}
hold_raw$purpose <- as.factor(hold_raw$purpose)
```

##### Initial List Status

```{r data preprocessing}
hold_raw$initial_list_status <- as.factor(hold_raw$initial_list_status)
```

##### Application Type

```{r data preprocessing}
hold_raw$application_type <- as.factor(hold_raw$application_type)
```

##### Hardship Flag

```{r data preprocessing}
hold_raw$hardship_flag <- as.factor(hold_raw$hardship_flag)
```

##### Debt Settlement Flag

```{r data preprocessing}
hold_raw$debt_settlement_flag <- as.factor(hold_raw$debt_settlement_flag)
```

#### Character-to-Numeric Conversion

##### Term

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    term= readr::parse_number(as.character(term))
  )
```

#### Character-to-Date Conversion

##### Issue Date

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    issue_d = lubridate::my(issue_d)
  )
```

##### Earliest Credit Line

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    earliest_cr_line = lubridate::my(earliest_cr_line)
  )
```

##### Last Credit Pull Date

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    last_credit_pull_d = lubridate::my(last_credit_pull_d)
  )
```

#### Missing Value Treatment

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    no_delinq_history_flag = ifelse(is.na(mths_since_last_delinq), 1, 0),
    mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 999, mths_since_last_delinq))

num_cols_hold <- names(hold_raw)[sapply(hold_raw, is.numeric)]

train_medians <- sapply(train_raw[num_cols_train], median, na.rm = TRUE)

for (v in num_cols_hold) {
  hold_raw[[v]][!is.finite(hold_raw[[v]])] <- NA_real_
  hold_raw[[v]][is.na(hold_raw[[v]])] <- train_medians[[v]]
}
```

#### Handling Rare or Sparse Categories

##### Application Type

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    application_type = forcats::fct_collapse(
      factor(application_type),
      OTHER = c("JOINT", "DIRECT_PAY")
    )
  )
```

##### Home Ownership

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    home_ownership = forcats::fct_collapse(
      factor(home_ownership),
      OTHER = c("OTHER", "ANY", "NONE")
    )
  )
```

#### Removing Unnecessary Variables

```{r data preprocessing}
hold_raw <- hold_raw %>%
    select(
      -emp_title,
      -title
    )
```

## *Feature Engineering*

### Ordinal Risk Features
_____________________________________________________________________________________________________________

The variables `grade_score` and `sub_grade_score` translate lender risk grades into ordered numeric scores. This is useful because grade A is lower risk than grade B, grade B is lower risk than grade C, and so on.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    grade_score = match(stringr::str_to_upper(stringr::str_trim(grade)), LETTERS[1:7]),
    sub_grade_score = as.integer(factor(
      sub_grade,
      levels = as.vector(outer(LETTERS[1:7], 1:5, paste0))
    ))
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    grade_score = match(stringr::str_to_upper(stringr::str_trim(grade)), LETTERS[1:7]),
    sub_grade_score = as.integer(factor(
      sub_grade,
      levels = as.vector(outer(LETTERS[1:7], 1:5, paste0))
    ))
  )
```

### Sparse Count Binning
_____________________________________________________________________________________________________________

Sparse count variables such as public records, bankruptcies, recent inquiries, and delinquencies are grouped into categories. The bins `0`, `1`, and `2+` are easier to interpret and more stable than rare exact counts.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    pub_rec_group = factor(case_when(
      pub_rec == 0 ~ "0",
      pub_rec == 1 ~ "1",
      pub_rec >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    recent_inquiry_group = factor(case_when(
      is.na(inq_last_6mths) ~ "Missing",
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
     TRUE ~ "2+"
    )),
    any_pub_record = factor(ifelse(pub_rec > 0, "Yes", "No")),
    any_bankruptcy = factor(ifelse(pub_rec_bankruptcies > 0, "Yes", "No")),
    any_current_delinq = factor(ifelse(acc_now_delinq > 0, "Yes", "No"))
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    pub_rec_group = factor(case_when(
      pub_rec == 0 ~ "0",
      pub_rec == 1 ~ "1",
      pub_rec >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    recent_inquiry_group = factor(case_when(
      is.na(inq_last_6mths) ~ "Missing",
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
      TRUE ~ "2+"
    )),
    any_pub_record = factor(ifelse(pub_rec > 0, "Yes", "No")),
    any_bankruptcy = factor(ifelse(pub_rec_bankruptcies > 0, "Yes", "No")),
    any_current_delinq = factor(ifelse(acc_now_delinq > 0, "Yes", "No"))
  )
```

### Credit History Features
_____________________________________________________________________________________________________________

The variable `fico_score` combines the lower and upper FICO range bounds into an approximate average score. `credit_history_years` measures the length of credit history between the earliest credit line and loan issue date.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2,
    fico_range_width = fico_range_high - fico_range_low,
    issue_year = lubridate::year(issue_d),
    credit_history_days_at_issue = as.numeric(issue_d - earliest_cr_line),
    credit_history_years_at_issue = credit_history_days_at_issue / 365.25,
    last_pull_after_issue_days = as.numeric(last_credit_pull_d - issue_d)
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2,
    fico_range_width = fico_range_high - fico_range_low,
    issue_year = lubridate::year(issue_d),
    credit_history_days_at_issue = as.numeric(issue_d - earliest_cr_line),
    credit_history_years_at_issue = credit_history_days_at_issue / 365.25,
    last_pull_after_issue_days = as.numeric(last_credit_pull_d - issue_d)
  )
```

### Ratio Features
_____________________________________________________________________________________________________________

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    loan_to_income = loan_amnt / pmax(annual_inc, 1),
    installment_to_monthly_income = installment / pmax(annual_inc / 12, 1),
    installment_to_loanamt = installment / pmax(loan_amnt, 1),
    revol_bal_to_income = revol_bal / pmax(annual_inc, 1),
    total_credit_lines_per_open = total_acc / pmax(open_acc, 1)
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    loan_to_income = loan_amnt / pmax(annual_inc, 1),
    installment_to_monthly_income = installment / pmax(annual_inc / 12, 1),
    installment_to_loanamt = installment / pmax(loan_amnt, 1),
    revol_bal_to_income = revol_bal / pmax(annual_inc, 1),
    total_credit_lines_per_open = total_acc / pmax(open_acc, 1)
  )
```

### Creating ZIP Feature
_____________________________________________________________________________________________________________

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    zip_code = stringr::str_extract(address, "\\d{5}$")
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    zip_code = stringr::str_extract(address, "\\d{5}$")
  )
```

### Outlier Flags
_____________________________________________________________________________________________________________

```{r}
outlier_vars <- c(
  "annual_inc", "dti", "open_acc", "pub_rec",
  "revol_bal", "mort_acc", "loan_amnt"
)

outlier_caps <- purrr::map_dbl(
  outlier_vars,
  ~ quantile(train_raw[[.x]], probs = 0.99, na.rm = TRUE, names = FALSE)
)

names(outlier_caps) <- outlier_vars

for (v in outlier_vars) {
  cap_value <- outlier_caps[[v]]

  train_raw[[paste0(v, "_outlier_flag")]] <- ifelse(
    !is.na(train_raw[[v]]) & train_raw[[v]] > cap_value, 1, 0
  )

  hold_raw[[paste0(v, "_outlier_flag")]] <- ifelse(
    !is.na(hold_raw[[v]]) & hold_raw[[v]] > cap_value, 1, 0
  )

  train_raw[[v]] <- pmin(train_raw[[v]], cap_value)
  hold_raw[[v]] <- pmin(hold_raw[[v]], cap_value)
}

train_raw <- train_raw %>%
  mutate(
    log_annual_inc = log1p(pmax(annual_inc, 0)),
    log_revol_bal = log1p(pmax(revol_bal, 0)),
    log_loan_amnt = log1p(pmax(loan_amnt, 0))
  )

hold_raw <- hold_raw %>%
  mutate(
    log_annual_inc = log1p(pmax(annual_inc, 0)),
    log_revol_bal = log1p(pmax(revol_bal, 0)),
    log_loan_amnt = log1p(pmax(loan_amnt, 0))
  )
```


### Feature Removal
_____________________________________________________________________________________________________________

The cleaned dataset has no missing values after preprocessing and contains a mixture of numeric, categorical, and engineered predictors.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  select(
    -address,
    -issue_d,
    -earliest_cr_line,
    -last_credit_pull_d,
    -grade,
    -sub_grade,
    -fico_range_low,
    -fico_range_high,
    -installment,
    -annual_inc,
    -inq_last_6mths,
    -acc_now_delinq,
    -pub_rec_bankruptcies
  )
```


#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  select(
    -address,
    -issue_d,
    -earliest_cr_line,
    -last_credit_pull_d,
    -grade,
    -sub_grade,
    -fico_range_low,
    -fico_range_high,
    -installment,
    -annual_inc,
    -inq_last_6mths,
    -acc_now_delinq,
    -pub_rec_bankruptcies
  )
```

### Manual Rare Category Handling
____________________________________________________________________________________________________________

```{r data preprocessing}
collapse_levels <- function(train_col, hold_col, min_prop = 0.01) {
  train_chr <- as.character(train_col)

  keep_levels <- names(prop.table(table(train_chr, useNA = "no"))[
    prop.table(table(train_chr, useNA = "no")) >= min_prop
  ])

  train_out <- ifelse(
    is.na(train_chr), "Missing",
    ifelse(train_chr %in% keep_levels, train_chr, "Other")
  )

  hold_chr <- as.character(hold_col)

  hold_out <- ifelse(
    is.na(hold_chr), "Missing",
    ifelse(hold_chr %in% keep_levels, hold_chr, "Other")
  )

  final_levels <- sort(unique(c(train_out, "Missing", "Other")))

  list(
    train = factor(train_out, levels = final_levels),
    hold = factor(hold_out, levels = final_levels)
  )
}
```

```{r data preprocessing}
categorical_cols <- names(train_raw)[sapply(train_raw, function(x) is.character(x) || is.factor(x))]
categorical_cols <- setdiff(categorical_cols, "loan_default")

for (v in categorical_cols) {
  collapsed <- collapse_levels(train_raw[[v]], hold_raw[[v]], min_prop = 0.01)

  train_raw[[v]] <- collapsed$train
  hold_raw[[v]] <- collapsed$hold
}
```


```{r data preprocessing}
num_cols_train <- names(train_raw)[sapply(train_raw, is.numeric)]

train_medians <- sapply(train_raw[num_cols_train], median, na.rm = TRUE)
train_medians[is.na(train_medians)] <- 0

for (v in num_cols_train) {
  train_raw[[v]][!is.finite(train_raw[[v]])] <- NA_real_
  hold_raw[[v]][!is.finite(hold_raw[[v]])] <- NA_real_

  train_raw[[v]][is.na(train_raw[[v]])] <- train_medians[[v]]
  hold_raw[[v]][is.na(hold_raw[[v]])] <- train_medians[[v]]
}
```

## *Re-Assignment*
____________________________________________________________________________________________________________

Since the datasets are "cleaned", the `raw` denotion is removed.

```{r assign}
# TRAINING DATASET RE-ASSIGNMENT
train <- train_raw

# HOLDOUT DATASET RRE-ASSIGNMENT
hold <- hold_raw
```

## *Dimension Reduction*

Dimension reduction is considered when a model uses many correlated or dummy-encoded predictors. PCA can reduce many numeric columns into a smaller set of components, but it also reduces interpretability. PCA is most useful for models that are sensitive to high-dimensional numeric input, such as neural networks or regularized models using a wide dummy matrix. It is less necessary for tree-based models because tree models can handle nonlinear splits and mixed predictor importance directly. PCA requires numeric predictors to be centered and scaled. Without scaling, variables measured on large numeric scales can dominate the components.

The code below corresponds to the appropriate PCA process. This code is not evaluated and, therefore, is optional. It is included for completeion.

```{r dimension reduction, eval=FALSE}
train_dummy <- fastDummies::dummy_cols(
  train,
  remove_first_dummy = TRUE,
  remove_selected_columns = TRUE
)
```

```{r dimension reduction, eval=FALSE}
hold_dummy <- fastDummies::dummy_cols(
  hold,
  remove_first_dummy = TRUE,
  remove_selected_columns = TRUE
)
```

```{r dimension reduction, eval=FALSE}
missing_hold_cols <- setdiff(names(train_dummy), names(hold_dummy))
for (v in missing_hold_cols) hold_dummy[[v]] <- 0
```

```{r dimension reduction, eval=FALSE}
extra_hold_cols <- setdiff(names(hold_dummy), names(train_dummy))
hold_dummy <- hold_dummy %>% select(-all_of(extra_hold_cols))
hold_dummy <- hold_dummy %>% select(all_of(names(train_dummy)))
```

```{r dimension reduction, eval=FALSE}
pca_recipe <- recipes::recipe(loan_default ~ ., data = train_dummy) %>%
  recipes::step_zv(all_predictors()) %>%
  recipes::step_normalize(all_numeric_predictors()) %>%
  recipes::step_pca(all_numeric_predictors(), threshold = 0.90)
```

```{r dimension reduction, eval=FALSE}
pca_prep <- recipes::prep(pca_recipe, training = train_dummy)
```
```{r dimension reduction, eval=FALSE}
train_pca <- recipes::bake(pca_prep, new_data = train_dummy)
```

```{r dimension reduction, eval=FALSE}
hold_pca <- recipes::bake(pca_prep, new_data = hold_dummy)
```

## *Sample*

The significance of the class imbalance can be exemplified by the following explanation. If a model predicts most loans as non-default, it can obtain high accuracy while failing to identify default cases. Because defaults are the minority class, sampling and threshold tuning are considered as part of the modeling workflow.

### Class Distribution in Training Dataset
_____________________________________________________________________________________________________________

The target variable distribution observed shows an imbalanced, which also matches previous observations found from the target variable distributions.

```{r sample}
table(train$loan_default)
```

### Class Distribution in Holdout Dataset
_____________________________________________________________________________________________________________

For consistency purposes, the target variable distribution of the holdout dataset was also examined. A similar imbalance was observed as found in the training dataset target variable distriibution.

```{r sample}
table(hold$loan_default)
```

### Sampling Strategy
_____________________________________________________________________________________________________________

The main sampling option considered is undersampling the majority class. Undersampling creates a balanced training dataset by keeping all default cases and sampling an equal number of non-default cases. This can increase recall for the default class, but it may discard useful information from the majority class.

The model search compares full-distribution training with undersampled training. The current highest-F1 model was trained on the full training distribution, so the final selected model does not require undersampling.

```{r sample}
set.seed(123)

maj <- train %>%
  filter(loan_default == "No")
min <- train %>%
  filter(loan_default == "Yes")

maj_under <- maj %>%
  sample_n(nrow(min))

train_under <- bind_rows(min, maj_under) %>%
  slice_sample(prop = 1)

prop.table(table(train_under$loan_default))
```

The holdout dataset is not undersampled. It remains in its natural distribution so final evaluation reflects the type of class balance expected on new loan records.

```{r sample}
prop.table(table(hold$loan_default))
```

## *Exporting Cleaned Data For Modeling*

### Cleaned Training Dataset
_____________________________________________________________________________________________________________

```{r clean export}
# IMBALANCED TRAINING SET
saveRDS(train, file.path(project_dir, 'train.rds'))

# UNDERSAMPLED TRAINING SET
saveRDS(train_under, file.path(project_dir, 'train_under.rds'))
```

### Cleaned Holdout Dataset
_____________________________________________________________________________________________________________

```{r clean export}
saveRDS(hold, file.path(project_dir, 'holdout.rds'))
```

The procedures for producing the discussed models can be located in the `models` folder. The corresponding `html` and `rmd` model files can be found. The `html` files are the resulting files with output. While, the `rmd` files are the r chunks and the inputs with no outputs.

Generally, the modeling workflow was as follows:

1. Introduction
2. Read in the Data
3. Preview the Data
4. Specialized Pre-Model Processing (if necessary)
5. Set Up Model Parameters
6. Build Model & Train on Training Dataset
7. View Model Summary
8. Predict on Holdout Dataset
9. Create Table of Class and Probability Predictions
10. Save Model & Results

The models analyzed were logistic regression, decision trees, random forest ensembles, boosting ensembles, and neural networks. All models were segmented into full and balanced training datasets during the model construction.

---

# **Results**

## *Decision Tree*

### Derived Using Gini Impurity Measure
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
gini_tree_results <- readRDS(file.path(project_dir, "gini_tree_results.rds"))
print(gini_tree_results)
```

#### Undersampled Training Dataset

```{r}
gini_tree_us_results <- readRDS(file.path(project_dir, "gini_tree_us_results.rds"))
print(gini_tree_us_results)
```

### Derived Using Entropy Index Impurity Measure
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
s_tree_results <- readRDS(file.path(project_dir, "entropy_tree_results.rds"))
print(s_tree_results)
```

#### Undersampled Training Dataset

```{r}
s_tree_us_results <- readRDS(file.path(project_dir, "entropy_tree_us_results.rds"))
print(s_tree_us_results)
```

### Derived Using Prunning With Cross Validation - Gini Index Impurity Measure
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
pruned_cv_gini_tree_results <- readRDS(file.path(project_dir, "pruned_cv_gini_tree_results.rds"))
print(pruned_cv_gini_tree_results)
```

#### Undersampled Training Dataset

```{r}
pruned_cv_gini_tree_us_results <- readRDS(file.path(project_dir, "pruned_cv_gini_tree_us_results.rds"))
print(pruned_cv_gini_tree_us_results)
```

### Derived Using Prunning With Cross Validation - Entropy Index Impurity Measure
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
pruned_cv_entropy_tree_results <- readRDS(file.path(project_dir, "pruned_cv_entropy_tree_results.rds"))
print(pruned_cv_entropy_tree_results)
```

#### Undersampled Training Dataset

```{r}
pruned_cv_entropy_tree_us_results <- readRDS(file.path(project_dir, "pruned_cv_entropy_tree_results.rds"))
print(pruned_cv_entropy_tree_us_results)
```

## *Random Forest Ensembles*

### Derived Using randomForest - No Cross Validation
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
rf_results <- readRDS(file.path(project_dir, "rf_results.rds"))
print(rf_results)
```

#### Undersampled Training Dataset

```{r}
rf_us_results <- readRDS(file.path(project_dir, "rf_us_results.rds"))
print(rf_us_results)
```

### Derived Using Ranger - No Cross Validation
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
ranger_rf_results <- readRDS(file.path(project_dir, "ranger_rf_results.rds"))
print(ranger_rf_results)
```

#### Undersampled Training Dataset

```{r}
ranger_rf_us_results <- readRDS(file.path(project_dir, "ranger_rf_us_results.rds"))
print(ranger_rf_us_results)
```

### Derived Using randomForest - With Cross Validation
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
rf_cv_results <- readRDS(file.path(project_dir, "rf_cv_results.rds"))
print(rf_cv_results)
```

#### Undersampled Training Dataset

```{r}
rf_cv_us_results <- readRDS(file.path(project_dir, "rf_cv_us_results.rds"))
print(rf_cv_us_results)
```

### Derived Using Ranger - With Cross Validation
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
ranger_cv_results <- readRDS(file.path(project_dir, "ranger_cv_results.rds"))
print(ranger_cv_results)
```

#### Undersampled Training Dataset

```{r}
ranger_cv_us_results <- readRDS(file.path(project_dir, "ranger_cv_us_results.rds"))
print(ranger_cv_us_results)
```

## *Boosting Ensembles*

### Derived From Shallow XGBoost
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
xgb_shallow_results <- readRDS(file.path(project_dir, "xgb_shallow_results.rds"))
print(xgb_shallow_results)
```

#### Undersampled Training Dataset

```{r}
xgb_shallow_us_results <- readRDS(file.path(project_dir, "xgb_shallow_us_results.rds"))
print(xgb_shallow_us_results)
```

### Derived From Medium XGBoost
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
xgb_medium_results <- readRDS(file.path(project_dir, "xgb_medium_results.rds"))
print(xgb_medium_results)
```

#### Undersampled Training Dataset

```{r}
xgb_medium_us_results <- readRDS(file.path(project_dir, "xgb_medium_us_results.rds"))
print(xgb_medium_us_results)
```

### Derived From Slow XGBoost
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
xgb_slow_results <- readRDS(file.path(project_dir, "xgb_slow_results.rds"))
print(xgb_slow_results)
```

#### Undersampled Training Dataset

```{r}
xgb_slow_us_results <- readRDS(file.path(project_dir, "xgb_slow_us_results.rds"))
print(xgb_slow_us_results)
```

### Derived From LogitBoost 50 Iterations
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
logitboost_50_results <- readRDS(file.path(project_dir, "logitboost_50_results.rds"))
print(logitboost_50_results)
```

#### Undersampled Training Dataset

```{r}
logitboost_50_us_results <- readRDS(file.path(project_dir, "logitboost_50_us_results.rds"))
print(logitboost_50_us_results)
```

### Derived From LogitBoost 100 Iterations
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
logitboost_100_results <- readRDS(file.path(project_dir, "logitboost_100_results.rds"))
print(logitboost_100_results)
```

#### Undersampled Training Dataset

```{r}
logitboost_100_us_results <- readRDS(file.path(project_dir, "logitboost_100_us_results.rds"))
print(logitboost_100_us_results)
```

## *Logistic Regression*

### Derived From Lasso Regularization
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
log_lasso_results <- readRDS(file.path(project_dir, "log_lasso_results.rds"))
print(log_lasso_results)
```

#### Undersampled Training Dataset

```{r}
log_lasso_us_results <- readRDS(file.path(project_dir, "log_lasso_undersampled_results.rds"))
print(log_lasso_us_results)
```

### Derived From Ridge Regularization
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
log_ridge_results <- readRDS(file.path(project_dir, "log_ridge_results.rds"))
print(log_ridge_results)
```

#### Undersampled Training Dataset

```{r}
log_ridge_us_results <- readRDS(file.path(project_dir, "log_ridge_undersampled_results.rds"))
print(log_ridge_us_results)
```

### Derived From Elastic Net
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
log_elastic_results <- readRDS(file.path(project_dir, "log_elastic_results.rds"))
print(log_elastic_results)
```

#### Undersampled Training Dataset

```{r}
log_elastic_us_results <- readRDS(file.path(project_dir, "log_elastic_undersampled_results.rds"))
print(log_elastic_us_results)
```

## *Neural Network*

### Derived From Lasso Selection
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
nn_logistic_results <- readRDS(file.path(project_dir, "nn_logistic_results.rds"))
print(nn_logistic_results)
```

#### Undersampled Training Dataset

```{r}
nn_logistic_us_results <- readRDS(file.path(project_dir, "nn_logistic_us_results.rds"))
print(nn_logistic_us_results)
```

### Derived From Random Forest Variable Importance
____________________________________________________________________________________________________________

#### Training Dataset

```{r}
nn_rf_results <- readRDS(file.path(project_dir, "nn_rf_results.rds"))
print(nn_rf_results)
```

#### Undersampled Training Dataset

```{r}
nn_rf_us_results <- readRDS(file.path(project_dir, "nn_rf_us_results.rds"))
print(nn_rf_us_results)
```

---

# **Discussion**

The models showed a consistent tradeoff between performance on the full training dataset and performance on the undersampled training dataset. In general, models trained on the full dataset tended to produce higher accuracy, precision, and specificity, meaning they were better at correctly identifying non-default cases and avoiding false positives. However, the undersampled models usually produced higher sensitivity/recall and balanced accuracy, meaning they were better at detecting loan defaults. This pattern is expected in an imbalanced classification problem because the full dataset contains many more non-default observations, while undersampling forces the model to learn more directly from the minority default class.

For the decision tree models, the full training dataset produced stronger F1 scores, while the undersampled training dataset improved recall and balanced accuracy. The full Gini decision tree achieved an accuracy of 0.916, precision of 0.850, sensitivity of 0.697, specificity of 0.970, F1 score of 0.766, and AUC of 0.925. The undersampled Gini tree lowered accuracy to 0.878 and precision to 0.640, but sensitivity increased sharply to 0.865 and balanced accuracy improved from 0.833 to 0.873. The same pattern appeared for the entropy tree: the full model had a higher F1 score of 0.766, while the undersampled model improved sensitivity from 0.698 to 0.863 and AUC from 0.926 to 0.944. The cross-validated Gini tree followed this trend as well, with the undersampled version reaching the highest decision-tree balanced accuracy of 0.874 and sensitivity of 0.873, but with a lower F1 score of 0.732 compared with 0.766 for the full model. Overall, the decision trees trained on the full data were better if precision and F1 score were prioritized, while the undersampled trees were better if the primary goal was to identify as many default cases as possible. One note is that the cross-validated entropy undersampled result appears to load the same RDS file as the full cross-validated entropy model, so that comparison should be checked before making a final claim about that specific model.

For the random forest models, undersampling again improved recall but reduced precision and overall F1 score. The basic `randomForest` model trained on the full dataset achieved accuracy of 0.909, precision of 0.893, specificity of 0.982, and F1 score of 0.725, but sensitivity was only 0.610. The undersampled version increased sensitivity to 0.858 and balanced accuracy to 0.859, but precision dropped to 0.601 and F1 score declined slightly to 0.707. The `ranger` random forest showed nearly the same pattern: the full model had precision of 0.897 and specificity of 0.983, while the undersampled model had much higher sensitivity of 0.848 but lower precision of 0.597. Cross-validation improved the full random forest model, especially for F1 score. The cross-validated `randomForest` model trained on the full dataset had the best random forest F1 score at 0.750, with accuracy of 0.914, precision of 0.871, sensitivity of 0.659, specificity of 0.976, and AUC of 0.938. The cross-validated undersampled models still performed better on recall, but their lower precision kept their F1 scores below the full cross-validated model. Therefore, the best random forest model was the full-data cross-validated `randomForest` model.

The boosting models showed the strongest overall performance, especially the XGBoost models trained on undersampled data. The full XGBoost models were extremely conservative in predicting default. For example, the shallow XGBoost model trained on the full dataset had very high precision of 0.995 and specificity of 0.999, but sensitivity was only 0.479, meaning it missed many actual default cases. The undersampled shallow XGBoost model was more balanced, increasing sensitivity to 0.722 and F1 score to 0.743 while still keeping specificity high at 0.946. The medium XGBoost model was the strongest boosting model. Its full-data version had AUC of 0.951 and precision of 0.988, but sensitivity was only 0.503. After undersampling, the medium XGBoost model improved sensitivity to 0.757, balanced accuracy to 0.854, and F1 score to 0.773, while maintaining strong accuracy of 0.913, precision of 0.791, specificity of 0.951, and AUC of 0.952. The slow XGBoost model also benefited from undersampling, increasing F1 score from 0.644 to 0.754 and sensitivity from 0.476 to 0.766. However, it did not outperform the medium XGBoost model. The LogitBoost models were weaker overall. The full LogitBoost 100 model had sensitivity of 0.755 but only an F1 score of 0.663 and AUC of 0.891, while the undersampled LogitBoost 100 model had very high precision and specificity but very poor sensitivity of 0.331. Overall, the best boosting model was the undersampled medium XGBoost model.
The logistic regression models were stable but less competitive than the tree-based ensemble methods. Lasso, ridge, and elastic net all produced similar results, and the undersampled versions were identical to the full versions in the reported output. The lasso model had accuracy of 0.897, precision of 0.930, sensitivity of 0.516, specificity of 0.991, F1 score of 0.664, and AUC of 0.923. Ridge regression was slightly more conservative, with higher precision of 0.962 and specificity of 0.995, but lower sensitivity of 0.497 and F1 score of 0.655. Elastic net was the best logistic regression model by F1 score, with accuracy of 0.897, precision of 0.926, sensitivity of 0.518, specificity of 0.990, F1 score of 0.665, and AUC of 0.923. These results suggest that the logistic models were good at identifying low-risk non-default cases, but they were not flexible enough to capture many of the default cases.

The neural network models performed well, especially the model using variables selected from lasso logistic regression. The full neural network with lasso-selected variables achieved accuracy of 0.913, balanced accuracy of 0.831, precision of 0.834, sensitivity of 0.695, specificity of 0.966, F1 score of 0.758, and AUC of 0.948. The undersampled version improved sensitivity to 0.845 and balanced accuracy to 0.855, but precision dropped to 0.605 and F1 score declined to 0.705. The neural network using random forest variable importance was weaker. Its full-data version had an F1 score of 0.682 and AUC of 0.916, while the undersampled version improved sensitivity to 0.824 but only reached an F1 score of 0.694. This suggests that the lasso-selected variables created a cleaner feature set for the neural network than the random-forest-selected variables. Overall, the best neural network was the full-data neural network using lasso logistic regression variable selection.
Comparing the best model from each class, the best decision tree model was the full-data entropy decision tree, with an F1 score of 0.766 and AUC of 0.926. The best random forest model was the full-data cross-validated randomForest model, with an F1 score of 0.750 and AUC of 0.938. The best boosting model was the undersampled medium XGBoost model, with an F1 score of 0.773 and AUC of 0.952. The best logistic regression model was elastic net, with an F1 score of 0.665 and AUC of 0.923. The best neural network was the full-data lasso-selected neural network, with an F1 score of 0.758 and AUC of 0.948. Based on these comparisons, the best overall model was the undersampled medium XGBoost model. It had the highest F1 score among all models, tied or exceeded the strongest AUC values, and provided the best balance between correctly identifying defaults and avoiding excessive false positives.

A likely reason the best model came from the XGBoost class is that loan default prediction contains complex, nonlinear relationships and interactions among borrower characteristics, loan attributes, and risk indicators. Logistic regression can only model these relationships in a mostly linear way, and a single decision tree can be too simple or unstable. Random forests improve stability by averaging many trees, but boosting can be especially effective because each new tree focuses on errors made by the previous trees. The undersampled medium XGBoost model likely performed well because undersampling gave the model more opportunity to learn the minority default class, while the medium-depth boosted trees were flexible enough to capture interactions without becoming as conservative as the full-data XGBoost models. This produced a model that maintained strong specificity and AUC while substantially improving sensitivity and F1 score, making it the strongest option for predicting loan default.

---

# **Conclusion**

Overall, the modeling results show that the best-performing approach for predicting loan default was the undersampled medium XGBoost model. This model provided the strongest overall balance between predictive accuracy, default detection, and false-positive control. While several full-data models achieved very high precision and specificity, they often missed a larger share of actual default cases. In contrast, the undersampled medium XGBoost model improved sensitivity while still maintaining strong accuracy, precision, specificity, F1 score, and AUC.

The results also show that undersampling was useful for models that needed better recognition of the minority default class. Across many model classes, undersampled models increased sensitivity and balanced accuracy, though often at the cost of lower precision. This tradeoff is important in loan default prediction because failing to identify a likely default can be more costly than incorrectly flagging a lower-risk applicant for additional review. For that reason, a model with a strong balance between sensitivity and precision is more useful than a model that only maximizes overall accuracy.

Based on the full comparison, the undersampled medium XGBoost model is recommended as the final model. Its performance suggests that boosted tree methods are well suited for this problem because they can capture nonlinear patterns, interactions, and risk signals that simpler models may miss. In a practical lending context, this model could support risk screening by identifying applicants with higher default likelihood while still preserving enough specificity to avoid excessive false alarms.