---
title: "Decision Tree Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Introduction**

Decision tree models are supervised machine learning algorithms used for both classification and regression problems. In classification settings, decision trees predict a categorical outcome by recursively partitioning the data into increasingly homogeneous subgroups based on predictor variables. Each split represents an explicit decision rule, making the model structure transparent and interpretable. This interpretability is especially useful in applied prediction problems, such as loan default classification, where understanding how borrower and loan characteristics contribute to predicted risk is important.

A decision tree is composed of internal nodes, branches, and terminal nodes. Internal nodes represent decision rules based on predictor variables, whereas branches represent the outcomes of those rules. Terminal nodes, also called leaf nodes, contain the final predicted class. In a binary loan default problem, each terminal node classifies observations as either default or non-default, typically according to the majority class within that node. Terminal nodes may also report the number of observations from each class, allowing the analyst to assess node purity and classification confidence.

Decision trees are commonly constructed using recursive partitioning. This process begins with the full training dataset and evaluates possible splits across all predictors. For numeric predictors, the algorithm considers possible cutoff values; for categorical predictors, it evaluates possible groupings of categories. The best split is selected according to an impurity criterion, such as the Gini index or entropy. A split is preferred when it produces child nodes that are more homogeneous than the parent node. This process is repeated recursively until a stopping rule is met, such as a maximum tree depth, a minimum number of observations per node, or insufficient improvement in node purity.

Several algorithms can be used to construct decision trees, including Classification and Regression Trees (CART), C4.5, and Chi-Squared Automatic Interaction Detection (CHAID). CART is commonly implemented in R through the rpart package and uses binary splits based on either the Gini index or information gain. In contrast, C4.5 may use multiway splits and information gain ratio, while CHAID relies on chi-squared tests of independence. In this analysis, CART-based classification trees were used because the response variable, `loan_default`, is categorical.

Although decision trees offer practical advantages, including ease of interpretation, low deployment cost, and robustness to outliers, they are prone to overfitting. A very deep tree can capture noise in the training data and perform poorly on unseen observations. To reduce this risk, tree complexity can be controlled during model training through parameters such as maximum depth, minimum split size, minimum terminal node size, and the complexity parameter. In addition, pruning can be used after an initial tree is grown to remove branches that contribute little predictive value. Cross-validation provides a systematic way to select the appropriate pruning level by estimating model performance across multiple training-validation partitions.

In this project, decision tree models were evaluated as interpretable baseline classifiers for predicting loan default. Both Gini-based and entropy-based splitting rules were considered, along with shallow, deeper, pruned, and cross-validated tree variants. Model performance was assessed using holdout predictions and classification metrics, including F1 score, accuracy, precision, recall, balanced accuracy, and area under the receiver operating characteristic curve. This approach allowed decision trees to be compared with more complex models while preserving a transparent benchmark for understanding borrower default risk.

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS("train.rds")
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS("train_under.rds")
```

## *Holdout Dataset*

```{r}
hold <- readRDS("holdout.rds")
```

# **Preview the Data**

## *Training Dataset*

```{r}
skimr::skim(train)
```

## *Undersampled Training Dataset*

```{r}
skimr::skim(train_under)
```

## *Holdout Dataset*

```{r}
skimr::skim(hold)
```

# **Default Tree - Gini Impurity Measure**

## *Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)

gini_ct = rpart(
  loan_default ~ .,
  data = train,
  method = "class",
  parms=list(split="gini"),
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0,
                        cp=0.0002465959))
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(gini_ct)
```

### Prune
____________________________________________________________________________________________________________

```{r}
gini_tree=prune(gini_ct, cp=0.0002465959)
```

### Print Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(gini_tree, extra=1, fallen.leaves=FALSE)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
gini_tree_prob <- predict(
  object = gini_tree,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
gini_tree_class <- predict(
  object = gini_tree,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
gini_tree_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(gini_tree_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(gini_tree_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(gini_tree_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
gini_tree_metrics <- evaluate_classifier(
  actual = gini_tree_predictions$actual,
  predicted = gini_tree_predictions$predicted,
  probability_yes = gini_tree_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(gini_tree, file.path(project_dir, "gini_tree_model.rds"))
```

#### Save Results

```{r}
gini_tree_results <- gini_tree_metrics %>%
  mutate(
    model_type = "Pruned Decision Tree ",
    impurity = "Gini Index",
    sample = "Full"
  )

saveRDS(gini_tree_results, file.path(project_dir, "gini_tree_results.rds"))
```

## *Undersampled Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
gini_ct_us = rpart(
  loan_default ~ .,
  data = train_under,
  method = "class",
  parms=list(split="gini"),
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0,
                        cp=0.0002090704)
  )
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(gini_ct_us)
```

### Prune
____________________________________________________________________________________________________________

```{r}
gini_tree_us=prune(gini_ct_us, cp=0.0002465959)
```

### Print Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(gini_tree_us, extra=1, fallen.leaves=FALSE)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
gini_tree_prob_us <- predict(
  object = gini_tree_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
gini_tree_class_us <- predict(
  object = gini_tree_us,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
gini_tree_predictions_us <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(gini_tree_class_us, levels = c("No", "Yes")),
  probability_yes = as.numeric(gini_tree_prob_us[, "Yes"])
)
```

#### Preview Table

```{r}
head(gini_tree_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
gini_tree_metrics_us <- evaluate_classifier(
  actual = gini_tree_predictions_us$actual,
  predicted = gini_tree_predictions_us$predicted,
  probability_yes = gini_tree_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(gini_tree_us, file.path(project_dir, "gini_tree_us_model.rds"))
```

#### Save Results

```{r}
gini_tree_results_us <- gini_tree_metrics_us %>%
  mutate(
    model_type = "Pruned Decision Tree",
    impurity = "Gini Index",
    sample = "Undersampled"
  )

saveRDS(gini_tree_us_results, file.path(project_dir, "gini_tree_us_results.rds"))
```


# **Default Tree - Entropy Index Impurity Measure**

## *Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
entropic_ct = rpart(
  loan_default ~ .,
  data = train,
  method = "class",
  parms=list(split="information"),
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0,
                        cp=0.0001977532)
  )
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(entropic_ct)
```

### Prune
____________________________________________________________________________________________________________

```{r}
entropic_tree=prune(entropic_ct, cp=0.0001977532)
```

### Print Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(entropic_tree, extra=1, fallen.leaves=FALSE)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
entropic_tree_prob <- predict(
  object = entropic_tree,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
entropic_tree_class <- predict(
  object = entropic_tree,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
entropic_tree_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(entropic_tree_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(entropic_tree_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(entropic_tree_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
entropy_tree_metrics <- evaluate_classifier(
  actual = entropic_tree_predictions$actual,
  predicted = entropic_tree_predictions$predicted,
  probability_yes = entropic_tree_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(entropic_tree, file.path(project_dir, "entropy_tree_model.rds"))
```

#### Save Results

```{r}
entropy_tree_results <- entropy_tree_metrics %>%
  mutate(
    model_type = "Pruned Decision Tree ",
    impurity = "Entropy Index",
    sample = "Full"
  )

saveRDS(entropy_tree_results, file.path(project_dir, "entropy_tree_results.rds"))
```

## *Undersampled Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
entropic_ct_us = rpart(
  loan_default ~ .,
  data = train_under,
  method = "class",
  parms=list(split="information"),
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0,
                        cp=0.0001977532)
  )
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(entropic_ct_us)
```

### Prune
____________________________________________________________________________________________________________

```{r}
entropic_tree_us=prune(entropic_ct_us, cp=0.0002465959)
```

### Print Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(entropic_tree_us, extra=1, fallen.leaves=FALSE)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
entropic_tree_prob_us <- predict(
  object = entropic_tree_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
entropic_tree_class_us <- predict(
  object = entropic_tree_us,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
entropic_tree_predictions_us <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(entropic_tree_class_us, levels = c("No", "Yes")),
  probability_yes = as.numeric(entropic_tree_prob_us[, "Yes"])
)
```

#### Preview Table

```{r}
head(entropic_tree_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
entropy_tree_metrics_us <- evaluate_classifier(
  actual = entropic_tree_predictions_us$actual,
  predicted = entropic_tree_predictions_us$predicted,
  probability_yes = entropic_tree_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(entropic_tree_us, file.path(project_dir, "entropy_tree_us_model.rds"))
```

#### Save Results

```{r}
entropy_tree_results_us <- entropy_tree_metrics_us %>%
  mutate(
    model_type = "Pruned Decision Tree ",
    impurity = "Entropy Index",
    sample = "Undersampled"
  )

saveRDS(entropy_tree_us_results, file.path(project_dir, "entropy_tree_us_results.rds"))
```

# **Prune Tree With Cross Validation - Gini Index Impurity Measure**

## *Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
cv_gini_ct=rpart(loan_default ~ ., 
                 data=train, 
                 method="class",
                 parms=list(split="gini"),
                 control=rpart.control(
                   maxdepth=30,
                   cp=0.0001977532, 
                   xval=15))
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(cv_gini_ct)
```

### Plot `cp` Values
____________________________________________________________________________________________________________

```{r}
plotcp(cv_gini_ct)
```

### Extract Best `cp`
____________________________________________________________________________________________________________

```{r}
best_cp_gini=cv_gini_ct$cptable[which.min(cv_gini_ct$cptable[,"xerror"]), "CP"]
```

### Print Best `cp`
____________________________________________________________________________________________________________

```{r}
print(best_cp_gini)
```

### Select Best Prune Tree
____________________________________________________________________________________________________________

```{r}
pruned_cv_gini_tree=prune(cv_gini_ct, cp=best_cp_gini)
```

### Print Best Prune Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(pruned_cv_gini_tree, extra=1, fallen.leaves=FALSE)
```

### Obtain Variable Importance
____________________________________________________________________________________________________________

```{r}
 print(pruned_cv_gini_tree$variable.importance)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
pruned_cv_gini_tree_prob <- predict(
  object = pruned_cv_gini_tree,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
pruned_cv_gini_tree_class <- predict(
  object = pruned_cv_gini_tree,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
pruned_cv_gini_tree_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(pruned_cv_gini_tree_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(pruned_cv_gini_tree_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(pruned_cv_gini_tree_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
pruned_cv_gini_tree_metrics <- evaluate_classifier(
  actual = pruned_cv_gini_tree_predictions$actual,
  predicted = pruned_cv_gini_tree_predictions$predicted,
  probability_yes = pruned_cv_gini_tree_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(pruned_cv_gini_tree, file.path(project_dir, "pruned_cv_gini_tree_model.rds"))
```

#### Save Results

```{r}
pruned_cv_gini_tree_results_us <- pruned_cv_gini_tree_metrics %>%
  mutate(
    model_type = "Cross-Validated Pruned Decision Tree",
    impurity = "Gini Index",
    sample = "Full"
  )

saveRDS(pruned_cv_gini_results, file.path(project_dir, "pruned_cv_gini_results.rds"))
```

## *Undersampled Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
cv_gini_ct_us=rpart(loan_default ~ ., 
                 data=train_under, 
                 method="class",
                 parms=list(split="gini"),
                 control=rpart.control(
                   maxdepth=30,
                   cp=0.0001977532, 
                   xval=15))
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(cv_gini_ct_us)
```

### Plot `cp` Values
____________________________________________________________________________________________________________

```{r}
plotcp(cv_gini_ct_us)
```

### Extract Best `cp`
____________________________________________________________________________________________________________

```{r}
best_cp_gini_us=cv_gini_ct_us$cptable[which.min(cv_gini_ct_us$cptable[,"xerror"]), "CP"]
```

### Print Best `cp`
____________________________________________________________________________________________________________

```{r}
print(best_cp_gini_us)
```

### Select Best Prune Tree
____________________________________________________________________________________________________________

```{r}
pruned_cv_gini_tree_us=prune(cv_gini_ct_us, cp=best_cp_gini_us)
```

### Print Best Prune Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(pruned_cv_gini_tree_us, extra=1, fallen.leaves=FALSE)
```

### Obtain Variable Importance
____________________________________________________________________________________________________________

```{r}
 print(pruned_cv_gini_tree_us$variable.importance)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
pruned_cv_gini_tree_us_prob <- predict(
  object = pruned_cv_gini_tree_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
pruned_cv_gini_tree_us_class <- predict(
  object = pruned_cv_gini_tree_us,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
pruned_cv_gini_tree_predictions_us <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(pruned_cv_gini_tree_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(pruned_cv_gini_tree_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(pruned_cv_gini_tree_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
pruned_cv_gini_tree_metrics_us <- evaluate_classifier(
  actual = pruned_cv_gini_tree_predictions_us$actual,
  predicted = pruned_cv_gini_tree_predictions_us$predicted,
  probability_yes = pruned_cv_gini_tree_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(pruned_cv_gini_tree_us, file.path(project_dir, "pruned_cv_gini_tree_us_model.rds"))
```

#### Save Results

```{r}
pruned_cv_gini_tree_results_us <- pruned_cv_gini_tree_metrics_us %>%
  mutate(
    model_type = "Cross-Validated Pruned Decision Tree",
    impurity = "Gini Index",
    sample = "Undersampled"
  )

saveRDS(pruned_cv_gini_tree_us_results, file.path(project_dir, "pruned_cv_gini_tree_us_results.rds"))
```

# **Prune Tree With Cross Validation - Entropy Index Impurity Measure**

## *Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
cv_entropy_ct=rpart(loan_default ~ ., 
                 data=train, 
                 method="class",
                 parms=list(split="information"),
                 control=rpart.control(cp=0.0002358743, xval=15)
                )
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(cv_entropy_ct)
```

### Plot `cp` Values
____________________________________________________________________________________________________________

```{r}
plotcp(cv_entropy_ct) # plot the cp results against model size
```

### Extract Best `cp`
____________________________________________________________________________________________________________

```{r}
best_cp_entropy=cv_entropy_ct$cptable[which.min(cv_entropy_ct$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
```

### Print Best `cp`
____________________________________________________________________________________________________________

```{r}
print(best_cp_entropy)
```

### Select Best Prune Tree
____________________________________________________________________________________________________________

```{r}
pruned_cv_entropy_tree=prune(cv_entropy_ct, cp=best_cp_entropy)
```

### Print Best Prune Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(pruned_cv_entropy_tree, extra=1, fallen.leaves=FALSE)
```

### Obtain Variable Importance
____________________________________________________________________________________________________________

```{r}
print(pruned_cv_entropy_tree$variable.importance)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
pruned_cv_entropy_tree_prob <- predict(
  object = pruned_cv_entropy_tree,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
pruned_cv_entropy_tree_class <- predict(
  object = pruned_cv_entropy_tree,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
pruned_cv_entropy_tree_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(pruned_cv_entropy_tree_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(pruned_cv_entropy_tree_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(pruned_cv_entropy_tree_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
pruned_cv_entropy_tree_metrics <- evaluate_classifier(
  actual = pruned_cv_entropy_tree_predictions$actual,
  predicted = pruned_cv_entropy_tree_predictions$predicted,
  probability_yes = pruned_cv_entropy_tree_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(pruned_cv_entropy_tree, file.path(project_dir, "pruned_cv_entropy_tree_model.rds"))
```

#### Save Results

```{r}
pruned_cv_entropy_tree_results <- pruned_cv_entropy_tree_metrics %>%
  mutate(
    model_type = "Cross-Validated Pruned Decision Tree",
    impurity = "Entropy Index",
    sample = "Full"
  )

saveRDS(pruned_cv_entropy_tree_results, file.path(project_dir, "pruned_cv_entropy_tree_results.rds"))
```

## *Undersampled Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)
cv_entropy_ct_us=rpart(loan_default ~ ., 
                 data=train_under, 
                 method="class",
                 parms=list(split="information"),
                 control=rpart.control(cp=0.0002358743, xval=15)
                )
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(cv_entropy_ct_us)
```

### Plot `cp` Values
____________________________________________________________________________________________________________

```{r}
plotcp(cv_entropy_ct_us) # plot the cp results against model size
```

### Extract Best `cp`
____________________________________________________________________________________________________________

```{r}
best_cp_entropy_us=cv_entropy_ct_us$cptable[which.min(cv_entropy_ct_us$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
```

### Print Best `cp`
____________________________________________________________________________________________________________

```{r}
print(best_cp_entropy_us)
```

### Select Best Prune Tree
____________________________________________________________________________________________________________

```{r}
pruned_tree_entropy_us=prune(cv_entropy_ct_us, cp=best_cp_entropy_us)
```

### Print Best Prune Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(pruned_tree_entropy_us, extra=1, fallen.leaves=FALSE)
```

### Obtain Variable Importance
____________________________________________________________________________________________________________

```{r}
print(pruned_cv_entropy_tree_us$variable.importance)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
pruned_cv_entropy_tree_tree_us_prob <- predict(
  object = pruned_cv_gini_tree_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
pruned_cv_entropy_tree_us_class <- predict(
  object =pruned_cv_entropy_tree_tree_us,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
pruned_cv_entropy_tree_predictions_us <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(pruned_cv_entropy_tree_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(pruned_cv_entropy_tree_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(pruned_cv_entropy_tree_predictions_us)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
pruned_cv_entropy_tree_metrics_us <- evaluate_classifier(
  actual = pruned_cv_entropy_tree_predictions_us$actual,
  predicted = pruned_cv_entropy_tree_predictions_us$predicted,
  probability_yes = pruned_cv_entropy_tree_predictions_us$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(pruned_cv_entropy_tree_us, file.path(project_dir, "pruned_cv_entropy_tree_us_model.rds"))
```

#### Save Results

```{r}
pruned_cv_entropy_tree_results_us <- pruned_cv_entropy_tree_metrics_us %>%
  mutate(
    model_type = "Cross-Validated Pruned Decision Tree",
    impurity = "Entropy Index",
    sample = "Undersampled"
  )

saveRDS(pruned_cv_entropu_tree_us_results, file.path(project_dir, "pruned_cv_entropy_tree_us_results.rds"))
```