---
title: "Decision Tree Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Introduction**

Decision tree models are supervised machine learning algorithms used for both classification and regression problems. In classification settings, decision trees predict a categorical outcome by recursively partitioning the data into increasingly homogeneous subgroups based on predictor variables. Each split represents an explicit decision rule, making the model structure transparent and interpretable. This interpretability is especially useful in applied prediction problems, such as loan default classification, where understanding how borrower and loan characteristics contribute to predicted risk is important.

A decision tree is composed of internal nodes, branches, and terminal nodes. Internal nodes represent decision rules based on predictor variables, whereas branches represent the outcomes of those rules. Terminal nodes, also called leaf nodes, contain the final predicted class. In a binary loan default problem, each terminal node classifies observations as either default or non-default, typically according to the majority class within that node. Terminal nodes may also report the number of observations from each class, allowing the analyst to assess node purity and classification confidence.

Decision trees are commonly constructed using recursive partitioning. This process begins with the full training dataset and evaluates possible splits across all predictors. For numeric predictors, the algorithm considers possible cutoff values; for categorical predictors, it evaluates possible groupings of categories. The best split is selected according to an impurity criterion, such as the Gini index or entropy. A split is preferred when it produces child nodes that are more homogeneous than the parent node. This process is repeated recursively until a stopping rule is met, such as a maximum tree depth, a minimum number of observations per node, or insufficient improvement in node purity.

Several algorithms can be used to construct decision trees, including Classification and Regression Trees (CART), C4.5, and Chi-Squared Automatic Interaction Detection (CHAID). CART is commonly implemented in R through the rpart package and uses binary splits based on either the Gini index or information gain. In contrast, C4.5 may use multiway splits and information gain ratio, while CHAID relies on chi-squared tests of independence. In this analysis, CART-based classification trees were used because the response variable, `loan_default`, is categorical.

Although decision trees offer practical advantages, including ease of interpretation, low deployment cost, and robustness to outliers, they are prone to overfitting. A very deep tree can capture noise in the training data and perform poorly on unseen observations. To reduce this risk, tree complexity can be controlled during model training through parameters such as maximum depth, minimum split size, minimum terminal node size, and the complexity parameter. In addition, pruning can be used after an initial tree is grown to remove branches that contribute little predictive value. Cross-validation provides a systematic way to select the appropriate pruning level by estimating model performance across multiple training-validation partitions.

In this project, decision tree models were evaluated as interpretable baseline classifiers for predicting loan default. Both Gini-based and entropy-based splitting rules were considered, along with shallow, deeper, pruned, and cross-validated tree variants. Model performance was assessed using holdout predictions and classification metrics, including F1 score, accuracy, precision, recall, balanced accuracy, and area under the receiver operating characteristic curve. This approach allowed decision trees to be compared with more complex models while preserving a transparent benchmark for understanding borrower default risk.

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS("train.rds")
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS("train_under.rds")
```

## *Holdout Dataset*

```{r}
hold <- readRDS("holdout.rds")
```

# **Preview the Data**

## *Training Dataset*

```{r}
skimr::skim(train)
```

## *Undersampled Training Dataset*

```{r}
skimr::skim(train_under)
```

## *Holdout Dataset*

```{r}
skimr::skim(hold)
```

# **Default Tree - Gini Impurity Measure**

## *Training Dataset*

### Build Tree
____________________________________________________________________________________________________________

```{r}
set.seed(123)

gini_ct = rpart(
  loan_default ~ .,
  data = train,
  method = "class",
  parms=list(split="gini"),
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0,
                        cp=0.0002465959))
```

### Print `cp`
____________________________________________________________________________________________________________

```{r}
printcp(gini_ct)
```

### Prune
____________________________________________________________________________________________________________

```{r}
gini_tree=prune(gini_ct, cp=0.0002465959)
```

### Print Tree
____________________________________________________________________________________________________________

```{r}
rpart.plot(gini_tree, extra=1, fallen.leaves=FALSE)
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
gini_tree_prob <- predict(
  object = gini_tree,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
gini_tree_class <- predict(
  object = gini_tree,
  newdata = hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
gini_tree_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(gini_tree_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(gini_tree_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(gini_tree_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
gini_tree_metrics <- evaluate_classifier(
  actual = gini_tree_predictions$actual,
  predicted = gini_tree_predictions$predicted,
  probability_yes = gini_tree_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(gini_tree, file.path(project_dir, "gini_tree_model.rds"))
```

#### Save Results

```{r}
gini_tree_results <- gini_tree_metrics %>%
  mutate(
    model_type = "Decision Tree ",
    impurity_measure = "Gini Index",
    sample = "Full"
  )

saveRDS(gini_tree_results, file.path(project_dir, "gini_tree_results.rds"))
```
**Default Tree - Gini Index Impurity Measure**

*Build Tree*
The tree built below is built off of the impurity measure gini index. Futher explanation can be elaborate via the use of stopping criteria synonymous to each tree created without using cross validation. In other words, holding all building elements constant, what is the optimal tree without overfitting we can build with just changing the split impurity method. As stated, the tree below utilizes the impurity measure of gini index. 
```{r}
set.seed(123)
gini.ct = rpart(
  loan_default ~ ., # target variable
  data = ctrain, # data source for training default tree
  method = "class", #classification tree
  parms=list(split="gini"), # Impurity measure = gini
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0, # no cross validation
                        cp=0.0002465959)
  )
#saveRDS(gini.ct, "st_im_gini_clean_1.rds")
printcp(gini.ct)
```
*Prune and Print*
Since we have the best cp value, we can print our tree. We will use the prune() function to "trim" the tree based on the optimal cp value we found. Next, we will print the tree. The resulting printed tree below is the result of our pruning!
```{r}
gini_tree=prune(gini.ct, cp=0.0002465959)
rpart.plot(gini_tree, extra=1, fallen.leaves=FALSE)
```

*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.

```{r}
# print(gini_tree$variable.importance)
```
Interestingly, the top five most important variables were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico_score.

*Print Tree Rules*
Here, we print the rules used for each split in our "gini-controlled" tree. 
```{r}
# rpart.rules(gini.ct, 
#              style="tall"
#             )
```
Analysis of the tree performance will be discussed in the concluding section where we report our results of all trees using a confusion matrix and other analytical tools to measure the our default tree performances. 

**Default Tree - Entropy Index Impurity Measure**
*Build Tree*
The tree built below is built off of the impurity measure entropy. Futher explanation can be elaborate via the use of stopping criteria synonymous to each tree created without using cross validation. In other words, holding all building elements constant, what is the optimal tree without overfitting we can build with just changing the split impurity method. As stated, the tree below utilizes an impurity measure of entropy index. 
```{r}
set.seed(123)
entropic.ct = rpart(
  loan_default ~ ., # target variable
  data = ctrain, # data source for training default tree
  method = "class", #classification tree
  parms=list(split="information"), # sets impurity measure to entropy 
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0, # no cross validation
                        cp=0.0001977532)
  )
saveRDS(entropic.ct, "st_im_s_clean_2.rds")
printcp(entropic.ct)
```
*Prune and Print*
Since we have the best cp value, we can print our tree. We will use the prune() function to "trim" the tree based on the optimal cp value we found. Next, we will print the tree. The resulting printed tree below is the result of our pruning! We will call the entropy-based tree 'S_tree' because S represents the state function of entropy in physical science terms. 
```{r}
S_tree=prune(entropic.ct, cp=0.0001977532)
rpart.plot(S_tree, extra=1, fallen.leaves=FALSE)
```

*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.

```{r}
# print(S_tree$variable.importance)
```

Interestingly, the top five most important variables were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico_score, the same variables as for gini index. 

*Print Tree Rules*
Here, we print the rules used for each split in our "gini-controlled" tree. 
```{r}
# rpart.rules(entropic.ct, 
#              style="tall"
#             )
```
Analysis of the tree performance will be discussed in the concluding section where we report our results of all trees using a confusion matrix and other analytical tools to measure the our default tree performances.

**Prune Tree with Cross Validation**
*Build Tree with Cross Validation*
We did prune the trees above, but that was to the optimal cp found using cross validation here. The impurity measure used here in the cross-validated tree below is gini by default. We cross validated on 10 folds. We printed the complexity parameter table using the printcp() function. 
```{r}
set.seed(123)
cv_tree=rpart(loan_default ~ ., 
                 data=ctrain, 
                 method="class",
                 parms=list(split="gini"),
                 control=rpart.control(
                   maxdepth=30,
                   cp=0.0001977532, 
                                       xval=15)
                )
# saveRDS(cv_tree, "st_im_gini_cv_clean_2.rds")
printcp(cv_tree) # print the cp results for each model size
```

```{r}
set.seed(123)
scv_tree=rpart(loan_default ~ ., 
                 data=ctrain, 
                 method="class",
                 parms=list(split="information"),
                 control=rpart.control(cp=0.0002358743, xval=15)
                )
saveRDS(scv_tree, "st_im_s_cv_clean_2.rds")
printcp(scv_tree) # print the cp results for each model size
```

*Plot cp Values*

Here, we are plotting cp results to eventually get the best cp value. 
```{r}
plotcp(cv_tree) # plot the cp results against model size
```

```{r}
plotcp(scv_tree) # plot the cp results against model size
```
*Extract Best cp*
The code below extracts the best cp value. 
```{r}
best_cp=cv_tree$cptable[which.min(cv_tree$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
print(best_cp)
```

```{r}
sbest_cp=scv_tree$cptable[which.min(scv_tree$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
print(sbest_cp)
```

*Select and Print Best Pruned Tree*
Once we have the best cp value, we can select the best pruned tree and show it using the rpart.plot() function. 
```{r}
pruned_tree=prune(cv_tree, cp=best_cp)
rpart.plot(pruned_tree, extra=1, fallen.leaves=FALSE)
```

```{r}
spruned_tree=prune(scv_tree, cp=sbest_cp)
rpart.plot(spruned_tree, extra=1, fallen.leaves=FALSE)
```
*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.

```{r}
 print(pruned_tree$variable.importance)
```

```{r}
print(spruned_tree$variable.importance)
```

Higher importance values indicate greater influence on the model's predictions. The five most important variables for our pruned tree were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico score. 

# Building Decision Tree with a Balanced Training Dataset 

> We first need to undersample the majority class in order to get a 50/50 ratio of Yes and No for loan default. 

Here, we undersample the data, such the the number for each class is equivalent.
```{r}
set.seed(123)
maj <- ctrain %>% dplyr::filter(loan_default == "No")
min <- ctrain %>% dplyr::filter(loan_default == "Yes")
maj_under <- maj %>% dplyr::sample_n(nrow(min))
train_us <- dplyr::bind_rows(min, maj_under) %>% dplyr::slice_sample(prop = 1)
prop.table(table(train_us$loan_default))
```
**Default Tree - Gini Index Impurity Measure**
*Build Tree*
The tree built below is built off of the impurity measure gini index. Futher explanation can be elaborate via the use of stopping criteria synonymous to each tree created without using cross validation. In other words, holding all building elements constant, what is the optimal tree without overfitting we can build with just changing the split impurity method. As stated, the tree below utilizes the impurity measure of gini index. 
```{r}
set.seed(123)
gini.ct.us = rpart(
  loan_default ~ ., # target variable
  data = train_us, # data source for training default tree
  method = "class", #classification tree
  parms=list(split="gini"), # Impurity measure = gini
  control=rpart.control(maxdepth=30, 
                        minbucket=5, 
                        xval=0, # no cross validation
                        cp=0.0002090704)
  )
saveRDS(gini.ct.us, "st_bal_gini_clean_2.rds")
printcp(gini.ct.us)
```
*Prune and Print*
Since we have the best cp value, we can print our tree. We will use the prune() function to "trim" the tree based on the optimal cp value we found. Next, we will print the tree. The resulting printed tree below is the result of our pruning!
```{r}
gini_tree_us=prune(gini.ct.us, cp=0.0002090704)
rpart.plot(gini_tree_us, extra=1, fallen.leaves=FALSE)
```

*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.

```{r}
print(gini_tree_us$variable.importance)
```
Interestingly, the top five most important variables were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico_score.

*Print Tree Rules*
Here, we print the rules used for each split in our "gini-controlled" tree. 
```{r}
# rpart.rules(gini.ct.us, 
#              style="tall"
#             )
```
Analysis of the tree performance will be discussed in the concluding section where we report our results of all trees using a confusion matrix and other analytical tools to measure the our default tree performances. 

**Default Tree - Entropy Index Impurity Measure**
*Build Tree*
The tree built below is built off of the impurity measure entropy. Futher explanation can be elaborate via the use of stopping criteria synonymous to each tree created without using cross validation. In other words, holding all building elements constant, what is the optimal tree without overfitting we can build with just changing the split impurity method. As stated, the tree below utilizes an impurity measure of entropy index. 
```{r}
set.seed(123)
entropic.ct.us = rpart(
  loan_default ~ ., # target variable
  data = train_us, # data source for training default tree
  method = "class", #classification tree
  parms=list(split="information"), # sets impurity measure to entropy 
  control=rpart.control(maxdepth=20, 
                        minbucket=5, 
                        xval=0, # no cross validation
                        cp=0.0001822665)
  )
saveRDS(entropic.ct.us, "st_bal_s_clean_2.rds")
printcp(entropic.ct.us)
```

*Prune and Print*
Since we have the best cp value, we can print our tree. We will use the prune() function to "trim" the tree based on the optimal cp value we found. Next, we will print the tree. The resulting printed tree below is the result of our pruning! We will call the entropy-based tree 'S_tree' because S represents the state function of entropy in physical science terms. 
```{r}
S_tree_us=prune(entropic.ct.us, cp=0.0002215789)
rpart.plot(S_tree_us, extra=1, fallen.leaves=FALSE)
```

*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.
```{r}
print(S_tree_us$variable.importance)
```

Interestingly, the top five most important variables were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico_score, the same variables as for gini index. 

*Print Tree Rules*
Here, we print the rules used for each split in our "gini-controlled" tree. 
```{r}
# rpart.rules(entropic.ct.us, 
#              style="tall"
#             )
```
Analysis of the tree performance will be discussed in the concluding section where we report our results of all trees using a confusion matrix and other analytical tools to measure the our default tree performances.

**Prune Tree with Cross Validation**
*Build Tree with Cross Validation*
We did prune the trees above, but that was to the optimal cp found using cross validation here. The impurity measure used here in the cross-validated tree below is gini by default. We cross validated on 10 folds. We printed the complexity parameter table using the printcp() function. 
```{r}
# gini
set.seed(123)
g_cv_tree_us=rpart(loan_default ~ ., 
                 data=train_us, 
                 method="class",
                 parms=list(split="gini"), # sets impurity measure to entropy 
                 control=rpart.control(maxdepth=30, 
                        minbucket=2, 
                        xval=10,
                        cp=0 # no cross validation
                 )
                )
saveRDS(g_cv_tree_us, "st_bal_gini_cv_clean_2.rds")
printcp(g_cv_tree_us) # print the cp results for each model size
```

```{r}
# entropy
set.seed(123)
s_cv_tree_us=rpart(loan_default ~ ., 
                 data=train_us, 
                 method="class",
                 parms=list(split="information"), # sets impurity measure to entropy 
                 control=rpart.control(maxdepth=30, 
                        minbucket=2, 
                        xval=10, 
                        cp=0.00034309 # no cross validation
                        )
                )
saveRDS(s_cv_tree_us, "st_bal_s_cv_clean_2.rds")
printcp(s_cv_tree_us) # print the cp results for each model size
```

*Plot cp Values*
Here, we are plotting cp results to eventually get the best cp value. 
```{r}
plotcp(g_cv_tree_us) # plot the cp results against model size
```

```{r}
plotcp(s_cv_tree_us)
```

*Extract Best cp*
The code below extracts the best cp value. 
```{r}
g_best_cp_us=g_cv_tree_us$cptable[which.min(g_cv_tree_us$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
print(g_best_cp_us)
```

```{r}
s_best_cp_us=s_cv_tree_us$cptable[which.min(s_cv_tree_us$cptable[,"xerror"]), "CP"] #pick the tree with the lowest cross validated error
print(s_best_cp_us)
```

*Select and Print Best Pruned Tree*
Once we have the best cp value, we can select the best pruned tree and show it using the rpart.plot() function. 
```{r}
g_pruned_tree_us=prune(g_cv_tree_us, cp=g_best_cp_us)
rpart.plot(g_pruned_tree_us, extra=1, fallen.leaves=FALSE)
```

```{r}
s_pruned_tree_us=prune(s_cv_tree_us, cp=s_best_cp_us)
rpart.plot(s_pruned_tree_us, extra=1, fallen.leaves=FALSE)
```

*Use Tree to Predict on Unseen Data*
We can keep the predicted class and probability for later use using the following code:
```{r}
ctrain <- ctrain %>%
  select(-loan_amnt)
chold <- chold 
st.preds <- data.frame(
    loan_default = chold$loan_default,
    
    im.g.prob = predict(
      gini_tree,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"], #probability of "Yes"
    im.s.prob = predict(
      S_tree,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"],
    us.gini.prob = predict(
      gini_tree_us,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"], #probability of "Yes"

    us.entropy.prob = predict(
      S_tree_us,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"], #probability of "Yes"

    g.cv.pruned.prob = predict(
      g_pruned_tree_us,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"], #probability of "Yes"

    s.cv.pruned.prob = predict(
      s_pruned_tree_us,
                                     chold,
                                     type ="prob"
                                     )[,"Yes"] #probability of "Yes"
)
```

```{r}
# # Create prediction objects
# # pred.im.log  <- prediction(st.preds$im.g.prob,  st.preds$loan_default)
# # pred.bal.log <- prediction(st.preds$im.s.prob, st.preds$loan_default)
# pred.im.log   <- prediction(st.preds$us.gini.prob,   st.preds$loan_default)
# pred.bal.log  <- prediction(st.preds$us.entropy.prob,  st.preds$loan_default)
# pred.im.rf   <- prediction(st.preds$g.cv.pruned.prob,   st.preds$loan_default)
# pred.bal.rf  <- prediction(st.preds$s.cv.pruned.prob,  st.preds$loan_default)
# 
# # ROC curve performance objects
# perf.nn_logistic    <- performance(pred.im.log,  "tpr", "fpr")
# perf.nn_logistic_us <- performance(pred.bal.log, "tpr", "fpr")
# perf.nn_rf          <- performance(pred.im.rf,   "tpr", "fpr")
# perf.nn_rf_us       <- performance(pred.bal.rf,  "tpr", "fpr")
# perf.nn_rf          <- performance(pred.im.rf,   "tpr", "fpr")
# perf.nn_rf_us       <- performance(pred.bal.rf,  "tpr", "fpr")
# 
# # AUC objects
# auc.nn_logistic    <- performance(pred.im.log,  "auc")@y.values[[1]]
# auc.nn_logistic_us <- performance(pred.bal.log, "auc")@y.values[[1]]
# auc.nn_rf          <- performance(pred.im.rf,   "auc")@y.values[[1]]
# auc.nn_rf_us       <- performance(pred.bal.rf,  "auc")@y.values[[1]]
# auc.nn_rf          <- performance(pred.im.rf,   "auc")@y.values[[1]]
# auc.nn_rf_us       <- performance(pred.bal.rf,  "auc")@y.values[[1]]
# 
```

```{r}
library(dplyr)
library(MLmetrics)

# =============================================================
# 1. Helper function: compute F1 at a given threshold
# =============================================================
f1_for_threshold <- function(prob, actual, threshold) {
  
  preds <- ifelse(prob > threshold, "Yes", "No")
  preds <- factor(preds, levels = c("Yes", "No"))
  actual <- factor(actual, levels = c("Yes", "No"))
  
  MLmetrics::F1_Score(y_pred = preds,
                      y_true = actual,
                      positive = "Yes")
}

# =============================================================
# 2. Create threshold grid
# =============================================================
thresholds <- seq(0.01, 0.99, by = 0.01)

# =============================================================
# 3. Compute F1 curves for all 6 models
# =============================================================

# Imbalanced Gini
f1_curve_gini_im <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$im.g.prob,
                   st.preds$loan_default, t)
})

# Imbalanced Entropy
f1_curve_entropy_im <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$im.s.prob,
                   st.preds$loan_default, t)
})

# Balanced Gini
f1_curve_gini_us <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$us.gini.prob,
                   st.preds$loan_default, t)
})

# Balanced Entropy
f1_curve_entropy_us <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$us.entropy.prob,
                   st.preds$loan_default, t)
})

# CV-pruned Gini
f1_curve_gini_cv <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$g.cv.pruned.prob,
                   st.preds$loan_default, t)
})

# CV-pruned Entropy
f1_curve_entropy_cv <- sapply(thresholds, function(t) {
  f1_for_threshold(st.preds$s.cv.pruned.prob,
                   st.preds$loan_default, t)
})


# =============================================================
# 4. Plot all F1 curves
# =============================================================
plot(thresholds, f1_curve_gini_im, type = "l", lwd = 2, col = "blue",
     xlab = "Threshold", ylab = "F1 Score",
     main = "F1 Score vs Threshold for All Tree Models")

lines(thresholds, f1_curve_entropy_im,  col = "green",  lwd = 2)
lines(thresholds, f1_curve_gini_us,     col = "red",    lwd = 2)
lines(thresholds, f1_curve_entropy_us,  col = "gold",   lwd = 2)
lines(thresholds, f1_curve_gini_cv,     col = "purple", lwd = 2)
lines(thresholds, f1_curve_entropy_cv,  col = "orange", lwd = 2)

legend("bottomright",
       legend = c("Imbalanced Gini",
                  "Imbalanced Entropy",
                  "Balanced Gini",
                  "Balanced Entropy",
                  "CV-Pruned Gini",
                  "CV-Pruned Entropy"),
       col = c("blue", "green", "red", "gold", "purple", "orange"),
       lwd = 2)


# =============================================================
# 5. Identify optimal thresholds for each model
# =============================================================
best_results <- data.frame(
  Model = c("Imbalanced Gini",
            "Imbalanced Entropy",
            "Balanced Gini",
            "Balanced Entropy",
            "CV-Pruned Gini",
            "CV-Pruned Entropy"),
  
  Best_Threshold = c(
    thresholds[which.max(f1_curve_gini_im)],
    thresholds[which.max(f1_curve_entropy_im)],
    thresholds[which.max(f1_curve_gini_us)],
    thresholds[which.max(f1_curve_entropy_us)],
    thresholds[which.max(f1_curve_gini_cv)],
    thresholds[which.max(f1_curve_entropy_cv)]
  ),
  
  Best_F1 = c(
    max(f1_curve_gini_im),
    max(f1_curve_entropy_im),
    max(f1_curve_gini_us),
    max(f1_curve_entropy_us),
    max(f1_curve_gini_cv),
    max(f1_curve_entropy_cv)
  )
)

# Print summary results
print(best_results)

```

*Obtain Variable Importance*
Variable importance measures how much each predictor contributes to the decision-making process in a decision tree model. In rpart variable importance is computed by summing the total improvement in node impurity (Gini index or entropy) across all splits where that variable is used, weighted by the number of observations for those nodes. The code below gives the variable importance for our pruned tree above.

```{r}
 print(g_pruned_tree_us$variable.importance)
```

```{r}
 print(s_pruned_tree_us$variable.importance)
```

Higher importance values indicate greater influence on the model's predictions. The five most important variables for our pruned tree were found to be sub_grade_score, debt_settlement_flag, installment_to_loanamnt, term, and fico score. 

# Comparing Performance of the Decision Trees


**ROC Curves**
```{r}
# # ============================================================
# # 1. Create ROCR prediction objects from your data frame
# # ============================================================
# # ----- Imbalanced Trees -----
# gini_pred_im <- ROCR::prediction(single.tree.preds$im.g.prob,
#                                  single.tree.preds$Actual_Class)
# gini_roc_im  <- ROCR::performance(gini_pred_im, "tpr", "fpr")
# 
# entropy_pred_im <- ROCR::prediction(single.tree.preds$im.s.prob,
#                                     single.tree.preds$Actual_Class)
# entropy_roc_im  <- ROCR::performance(entropy_pred_im, "tpr", "fpr")
# 
# # ----- Balanced (US) Trees -----
# gini_pred_us <- ROCR::prediction(single.tree.preds$us.gini.prob,
#                                  single.tree.preds$Actual_Class)
# gini_roc_us  <- ROCR::performance(gini_pred_us, "tpr", "fpr")
# 
# entropy_pred_us <- ROCR::prediction(single.tree.preds$us.entropy.prob,
#                                     single.tree.preds$Actual_Class)
# entropy_roc_us  <- ROCR::performance(entropy_pred_us, "tpr", "fpr")
# 
# # ----- Cross-Validated Pruned Trees -----
# gini_pred_cv <- ROCR::prediction(single.tree.preds$g.cv.pruned.prob,
#                                  single.tree.preds$Actual_Class)
# gini_roc_cv  <- ROCR::performance(gini_pred_cv, "tpr", "fpr")
# 
# entropy_pred_cv <- ROCR::prediction(single.tree.preds$s.cv.pruned.prob,
#                                     single.tree.preds$Actual_Class)
# entropy_roc_cv  <- ROCR::performance(entropy_pred_cv, "tpr", "fpr")
# 
# # ============================================================
# # 2. Plot all ROC curves
# # ============================================================
# 
# plot(gini_roc_im, col = "blue", lwd = 2,
#      main = "ROC Curves for All Tree Models")
# 
# plot(entropy_roc_im, col = "green",  lwd = 2, add = TRUE)
# plot(gini_roc_cv,   col = "violet", lwd = 2, add = TRUE)
# 
# plot(gini_roc_us,   col = "red",   lwd = 2, add = TRUE)
# plot(entropy_roc_us,col = "gold",  lwd = 2, add = TRUE)
# plot(entropy_roc_cv,col = "orange",lwd = 2, add = TRUE)
# 
# # Reference line
# abline(a = 0, b = 1, col = "black", lty = 2, lwd = 2)
# 
# # Legend
# legend("bottomright",
#        legend = c("Imbal Gini",
#                   "Imbal Entropy",
#                   "CV Pruned Gini",
#                   "Bal Gini",
#                   "Bal Entropy",
#                   "CV Pruned Entropy"),
#        col = c("blue", "green", "violet", "red", "gold", "orange"),
#        lwd = 2)
```

```{r}
# # Compute AUC values
# gini_auc_im      <- ROCR::performance(gini_pred_im, "auc")@y.values[[1]]
# entropy_auc_im   <- ROCR::performance(entropy_pred_im, "auc")@y.values[[1]]
# gini_auc_cv      <- ROCR::performance(gini_pred_cv, "auc")@y.values[[1]]
# 
# gini_auc_us      <- ROCR::performance(gini_pred_us, "auc")@y.values[[1]]
# entropy_auc_us   <- ROCR::performance(entropy_pred_us, "auc")@y.values[[1]]
# entropy_auc_cv   <- ROCR::performance(entropy_pred_cv, "auc")@y.values[[1]]
# 
# cat("Imbalanced AUCs:\n")
# cat("  Gini:    ", round(gini_auc_im, 4), "\n")
# cat("  Entropy: ", round(entropy_auc_im, 4), "\n")
# cat("  CV Gini: ", round(gini_auc_cv, 4), "\n\n")
# 
# cat("Balanced AUCs:\n")
# cat("  Gini:    ", round(gini_auc_us, 4), "\n")
# cat("  Entropy: ", round(entropy_auc_us, 4), "\n")
# cat("  CV Entropy:", round(entropy_auc_cv, 4), "\n")
```

The AUC differs a little from model to model, but it seems that the entropy balanced decision tree had the optimal AUC value. 

# Conclusion

The best decision tree was found to be that with the highest AUC. This would be the decision tree trained on entropy impurity measure. We have saved this tree below.

```{r}
# Save predicted values
saveRDS(single.tree.preds, "st.preds_clean_4.rds")
````