---
title: "Neural Network Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    caTools,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    glmnet,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    nnet,
    pacman,
    pROC,
    randomForest,
    ranger,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    xgboost,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Introduction**

Neural networks are flexible supervised learning models that can be used for classification and regression problems. In binary classification, a neural network estimates the probability that an observation belongs to a target class. For this project, the target class is loan default, where the model attempts to distinguish borrowers who defaulted from borrowers who did not default. Neural networks are useful in this setting because they can model complex, nonlinear relationships among borrower characteristics, loan terms, credit history, income, debt obligations, and other financial predictors without requiring the analyst to specify the exact functional form of those relationships in advance.

Unlike decision trees and random forests, neural networks require additional data formatting before model fitting. Numeric predictors should be scaled so that variables measured on larger scales do not dominate model training. Categorical predictors must also be converted into numeric dummy variables so they can be used as inputs in the network. These formatting steps are necessary because neural networks estimate weights across input variables, hidden-layer neurons, and output classes. In addition, the target variable must be coded carefully so that the positive class, `Yes`, corresponds to loan default. This ordering is important because the evaluation focuses on the model’s ability to identify borrowers who defaulted.

A neural network contains an input layer, one or more hidden layers, and an output layer. Each predictor contributes to the input layer, and each input is connected to hidden neurons through estimated weights. The hidden layer allows the model to learn nonlinear combinations of predictors. The output layer then converts these learned patterns into predicted class probabilities. In this analysis, single-hidden-layer neural network models were evaluated using the `nnet` package. The main model parameters were `size`, which controls the number of hidden-layer neurons, and `decay`,` which applies weight decay regularization. Larger networks can model more complex relationships, while stronger weight decay can reduce overfitting by penalizing large model weights.

Overfitting is a major concern in neural network modeling. Because neural networks estimate many parameters, they can fit the training data closely while performing poorly on unseen data. This risk is especially important when the dataset includes noisy predictors, sparse categories, or many dummy variables. To reduce overfitting, the modeling workflow considered regularization through weight decay, limited hidden-layer size, scaled numeric inputs, and variable selection. Variable selection is important because each retained predictor increases the number of input connections in the network. In this project, neural network variants were evaluated using selected predictors from the cleaned modeling workflow so that model complexity remained controlled.

Neural networks offer several advantages for loan default prediction. They can learn nonlinear interactions, accommodate both continuous and categorical information after formatting, and provide flexible probability estimates for classification. However, they also have important limitations. They require careful preprocessing, are sensitive to tuning choices, may be computationally expensive, and are less interpretable than logistic regression or decision tree models. This reduced interpretability is often described as the black-box problem because the internal weights of the model do not translate as directly into simple decision rules.

The neural network models were evaluated using F1 score as the primary performance metric because loan default is an imbalanced classification problem. Precision, sensitivity/recall, balanced accuracy, accuracy, specificity, and AUC were also used to characterize model performance. Sensitivity was especially important because failing to identify borrowers who default may carry substantial financial risk. Therefore, the selected neural network model should not only achieve a competitive F1 score but also maintain strong sensitivity for the default class.

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS(file.path(project_dir, "train.rds"))
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS(file.path(project_dir, "train_under.rds"))
```

## *Holdout Dataset*

```{r}
hold <- readRDS(file.path(project_dir, "holdout.rds"))
```

# **Preview the Data**

## *Training Dataset*

```{r}
skimr::skim(train)
```

## *Undersampled Training Dataset*

```{r}
skimr::skim(train_under)
```

## *Holdout Dataset*

```{r}
skimr::skim(hold)
```

# **Check Target Variable Distributions**

## *Training Dataset*

```{r}
table(train$loan_default)
prop.table(table(train$loan_default))
```

## *Undersampled Training Dataset*

```{r}
table(train_under$loan_default)
prop.table(table(train_under$loan_default))
```

## *Holdout Dataset*

```{r}
table(hold$loan_default)
prop.table(table(hold$loan_default))
```

# **Format Target Variable**

The response variable is temporarily ordered with `Yes` as the first level for `caret::twoClassSummary()` during model training. Prediction tables and evaluation still explicitly identify `Yes` as the positive class.

## *Training Dataset*

```{r}
train <- train %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

## *Undersampled Training Dataset*

```{r}
train_under <- train_under %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

## *Holdout Dataset*

```{r}
hold <- hold %>%
  mutate(loan_default = factor(loan_default, levels = c("Yes", "No")))
```

# **Create `evaluate_classifier()`**

```{r}
evaluate_classifier <- function(actual, predicted, probability_yes, positive_class = "Yes") {
  actual <- factor(actual, levels = c("No", "Yes"))
  predicted <- factor(predicted, levels = c("No", "Yes"))

  cm <- caret::confusionMatrix(
    data = predicted,
    reference = actual,
    positive = positive_class
  )

  roc_obj <- pROC::roc(
    response = actual,
    predictor = probability_yes,
    levels = c("No", "Yes"),
    direction = "<",
    quiet = TRUE
  )

  precision <- cm$byClass["Precision"]
  recall <- cm$byClass["Sensitivity"]

  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall / (precision + recall)
  )

  tibble::tibble(
    accuracy = as.numeric(cm$overall["Accuracy"]),
    balanced_accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    precision = as.numeric(precision),
    sensitivity_recall = as.numeric(recall),
    specificity = as.numeric(cm$byClass["Specificity"]),
    f1_score = as.numeric(f1),
    auc = as.numeric(pROC::auc(roc_obj))
  )
}
```

# **Additional Neural Network Formatting**

## *Load Saved Feature-Selection Models*

### Lasso Logistic Regression Models

```{r}
log_lasso <- readRDS(file.path(project_dir, "log_lasso_model.rds"))
log_lasso_us <- readRDS(file.path(project_dir, "log_lasso_undersampled_model.rds"))
```

### Random Forest Models

```{r}
rf_fit <- readRDS(file.path(project_dir, "rf_cv_model.rds"))
rf_fit_us <- readRDS(file.path(project_dir, "rf_cv_us_model.rds"))
```

## *Extract Variables from Lasso Logistic Regression*

### Training Dataset

```{r}
coefs_matrix <- coef(
  log_lasso$finalModel,
  s = log_lasso$bestTune$lambda
)

coefs <- as.vector(coefs_matrix)
nonzero_idx <- which(coefs != 0)

logit_names <- rownames(coefs_matrix)[nonzero_idx]
logit_names <- logit_names[logit_names != "(Intercept)"]

logit_names
```

### Undersampled Training Dataset

```{r}
coefs_matrix_us <- coef(
  log_lasso_us$finalModel,
  s = log_lasso_us$bestTune$lambda
)

coefs_us <- as.vector(coefs_matrix_us)
nonzero_idx_us <- which(coefs_us != 0)

logit_names_us <- rownames(coefs_matrix_us)[nonzero_idx_us]
logit_names_us <- logit_names_us[logit_names_us != "(Intercept)"]

logit_names_us
```

## *Extract Variables from Random Forest Importance*

### Training Dataset

```{r}
rf_importance <- caret::varImp(rf_fit, scale = FALSE)$importance %>%
  tibble::rownames_to_column("variable") %>%
  mutate(
    importance = rowMeans(across(where(is.numeric)), na.rm = TRUE)
  ) %>%
  arrange(desc(importance))

rf_names <- rf_importance %>%
  filter(importance > 10) %>%
  pull(variable)

rf_names
```

### Undersampled Training Dataset

```{r}
rf_importance_us <- caret::varImp(rf_fit_us, scale = FALSE)$importance %>%
  tibble::rownames_to_column("variable") %>%
  mutate(
    importance = rowMeans(across(where(is.numeric)), na.rm = TRUE)
  ) %>%
  arrange(desc(importance))

rf_names_us <- rf_importance_us %>%
  filter(importance > 10) %>%
  pull(variable)

rf_names_us
```

## *Create Neural Network Formatting Functions*

```{r}
scale_matrix_pair <- function(train_x, hold_x) {
  train_x[!is.finite(train_x)] <- 0
  hold_x[!is.finite(hold_x)] <- 0

  scaled_train <- scale(train_x)
  center_values <- attr(scaled_train, "scaled:center")
  scale_values <- attr(scaled_train, "scaled:scale")
  scale_values[scale_values == 0] <- 1

  scaled_hold <- scale(
    hold_x,
    center = center_values,
    scale = scale_values
  )

  scaled_train[!is.finite(scaled_train)] <- 0
  scaled_hold[!is.finite(scaled_hold)] <- 0

  list(
    train = as.data.frame(scaled_train),
    hold = as.data.frame(scaled_hold)
  )
}
```

```{r}
prepare_nn_data <- function(training_data,
                            holdout_data,
                            selected_columns,
                            target = "loan_default") {
  train_x <- model.matrix(
    stats::as.formula(paste(target, "~ . - 1")),
    data = training_data
  )

  hold_x <- model.matrix(
    stats::as.formula(paste(target, "~ . - 1")),
    data = holdout_data
  )

  missing_hold_cols <- setdiff(colnames(train_x), colnames(hold_x))

  for (v in missing_hold_cols) {
    hold_x <- cbind(hold_x, 0)
    colnames(hold_x)[ncol(hold_x)] <- v
  }

  extra_hold_cols <- setdiff(colnames(hold_x), colnames(train_x))

  if (length(extra_hold_cols) > 0) {
    hold_x <- hold_x[, !colnames(hold_x) %in% extra_hold_cols, drop = FALSE]
  }

  hold_x <- hold_x[, colnames(train_x), drop = FALSE]

  selected_columns <- intersect(selected_columns, colnames(train_x))

  if (length(selected_columns) == 0) {
    stop("No selected variables matched the neural network model matrix.")
  }

  train_x <- train_x[, selected_columns, drop = FALSE]
  hold_x <- hold_x[, selected_columns, drop = FALSE]

  scaled <- scale_matrix_pair(train_x, hold_x)

  train_ready <- scaled$train %>%
    mutate(loan_default = factor(training_data[[target]], levels = c("Yes", "No")))

  hold_ready <- scaled$hold %>%
    mutate(loan_default = factor(holdout_data[[target]], levels = c("Yes", "No")))

  list(
    train = train_ready,
    hold = hold_ready,
    selected_variables = selected_columns
  )
}
```

## *Create Neural Network Modeling Sets*

### Logistic Regression Variable Selection - Training Dataset

```{r}
log_nn_data <- prepare_nn_data(
  training_data = train,
  holdout_data = hold,
  selected_columns = logit_names
)
```

### Logistic Regression Variable Selection - Undersampled Training Dataset

```{r}
log_nn_data_us <- prepare_nn_data(
  training_data = train_under,
  holdout_data = hold,
  selected_columns = logit_names_us
)
```

### Random Forest Variable Selection - Training Dataset

```{r}
rf_nn_data <- prepare_nn_data(
  training_data = train,
  holdout_data = hold,
  selected_columns = rf_names
)
```

### Random Forest Variable Selection - Undersampled Training Dataset

```{r}
rf_nn_data_us <- prepare_nn_data(
  training_data = train_under,
  holdout_data = hold,
  selected_columns = rf_names_us
)
```

# **Neural Network Using Logistic Regression Variables**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
nn_logistic_control <- caret::trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

nn_logistic_grid <- expand.grid(
  size = c(3, 5),
  decay = c(0.05, 0.12)
)
```

### Build Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

nn_logistic <- caret::train(
  loan_default ~ .,
  data = log_nn_data$train,
  method = "nnet",
  metric = "ROC",
  linout = FALSE,
  tuneGrid = nn_logistic_grid,
  trace = FALSE,
  maxit = 100,
  trControl = nn_logistic_control
)
```

### View Summary
____________________________________________________________________________________________________________

```{r}
nn_logistic
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
nn_logistic_prob <- predict(
  object = nn_logistic,
  newdata = log_nn_data$hold,
  type = "prob"
)
```

#### Predict Class

```{r}
nn_logistic_class <- predict(
  object = nn_logistic,
  newdata = log_nn_data$hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
nn_logistic_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(nn_logistic_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(nn_logistic_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(nn_logistic_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
nn_logistic_metrics <- evaluate_classifier(
  actual = nn_logistic_predictions$actual,
  predicted = nn_logistic_predictions$predicted,
  probability_yes = nn_logistic_predictions$probability_yes
)

nn_logistic_metrics
```

### Store & Export
____________________________________________________________________________________________________________

```{r}
nn_logistic_results <- nn_logistic_metrics %>%
  mutate(
    model_type = "Neural Network",
    variable_selection = "Lasso Logistic Regression",
    sample = "Full",
    train_control = "10-fold cross-validation",
    size = nn_logistic$bestTune$size,
    decay = nn_logistic$bestTune$decay
  )

saveRDS(nn_logistic, file.path(project_dir, "nn_logistic_model.rds"))
saveRDS(nn_logistic_results, file.path(project_dir, "nn_logistic_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
nn_logistic_us_control <- caret::trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

nn_logistic_us_grid <- expand.grid(
  size = c(3, 5),
  decay = c(0.05, 0.12)
)
```

### Build Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

nn_logistic_us <- caret::train(
  loan_default ~ .,
  data = log_nn_data_us$train,
  method = "nnet",
  metric = "ROC",
  linout = FALSE,
  tuneGrid = nn_logistic_us_grid,
  trace = FALSE,
  maxit = 100,
  trControl = nn_logistic_us_control
)
```

### View Summary
____________________________________________________________________________________________________________

```{r}
nn_logistic_us
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
nn_logistic_us_prob <- predict(
  object = nn_logistic_us,
  newdata = log_nn_data_us$hold,
  type = "prob"
)
```

#### Predict Class

```{r}
nn_logistic_us_class <- predict(
  object = nn_logistic_us,
  newdata = log_nn_data_us$hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
nn_logistic_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(nn_logistic_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(nn_logistic_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(nn_logistic_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
nn_logistic_us_metrics <- evaluate_classifier(
  actual = nn_logistic_us_predictions$actual,
  predicted = nn_logistic_us_predictions$predicted,
  probability_yes = nn_logistic_us_predictions$probability_yes
)

nn_logistic_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

```{r}
nn_logistic_us_results <- nn_logistic_us_metrics %>%
  mutate(
    model_type = "Neural Network",
    variable_selection = "Lasso Logistic Regression",
    sample = "Undersampled",
    train_control = "10-fold cross-validation",
    size = nn_logistic_us$bestTune$size,
    decay = nn_logistic_us$bestTune$decay
  )

saveRDS(nn_logistic_us, file.path(project_dir, "nn_logistic_us_model.rds"))
saveRDS(nn_logistic_us_results, file.path(project_dir, "nn_logistic_us_results.rds"))
```

# **Neural Network Using Random Forest Variables**

## *Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
nn_rf_control <- caret::trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

nn_rf_grid <- expand.grid(
  size = c(3, 5),
  decay = c(0.05, 0.12)
)
```

### Build Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

nn_rf <- caret::train(
  loan_default ~ .,
  data = rf_nn_data$train,
  method = "nnet",
  metric = "ROC",
  linout = FALSE,
  tuneGrid = nn_rf_grid,
  trace = FALSE,
  maxit = 100,
  trControl = nn_rf_control
)
```

### View Summary
____________________________________________________________________________________________________________

```{r}
nn_rf
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
nn_rf_prob <- predict(
  object = nn_rf,
  newdata = rf_nn_data$hold,
  type = "prob"
)
```

#### Predict Class

```{r}
nn_rf_class <- predict(
  object = nn_rf,
  newdata = rf_nn_data$hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
nn_rf_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(nn_rf_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(nn_rf_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(nn_rf_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
nn_rf_metrics <- evaluate_classifier(
  actual = nn_rf_predictions$actual,
  predicted = nn_rf_predictions$predicted,
  probability_yes = nn_rf_predictions$probability_yes
)

nn_rf_metrics
```

### Store & Export
____________________________________________________________________________________________________________

```{r}
nn_rf_results <- nn_rf_metrics %>%
  mutate(
    model_type = "Neural Network",
    variable_selection = "Random Forest Importance",
    sample = "Full",
    train_control = "10-fold cross-validation",
    size = nn_rf$bestTune$size,
    decay = nn_rf$bestTune$decay
  )

saveRDS(nn_rf, file.path(project_dir, "nn_rf_model.rds"))
saveRDS(nn_rf_results, file.path(project_dir, "nn_rf_results.rds"))
```

## *Undersampled Training Dataset*

### Set Parameters
____________________________________________________________________________________________________________

```{r}
nn_rf_us_control <- caret::trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

nn_rf_us_grid <- expand.grid(
  size = c(3, 5),
  decay = c(0.05, 0.12)
)
```

### Build Model
____________________________________________________________________________________________________________

```{r}
set.seed(123)

nn_rf_us <- caret::train(
  loan_default ~ .,
  data = rf_nn_data_us$train,
  method = "nnet",
  metric = "ROC",
  linout = FALSE,
  tuneGrid = nn_rf_us_grid,
  trace = FALSE,
  maxit = 100,
  trControl = nn_rf_us_control
)
```

### View Summary
____________________________________________________________________________________________________________

```{r}
nn_rf_us
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
nn_rf_us_prob <- predict(
  object = nn_rf_us,
  newdata = rf_nn_data_us$hold,
  type = "prob"
)
```

#### Predict Class

```{r}
nn_rf_us_class <- predict(
  object = nn_rf_us,
  newdata = rf_nn_data_us$hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
nn_rf_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(nn_rf_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(nn_rf_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(nn_rf_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
nn_rf_us_metrics <- evaluate_classifier(
  actual = nn_rf_us_predictions$actual,
  predicted = nn_rf_us_predictions$predicted,
  probability_yes = nn_rf_us_predictions$probability_yes
)

nn_rf_us_metrics
```

### Store & Export
____________________________________________________________________________________________________________

```{r}
nn_rf_us_results <- nn_rf_us_metrics %>%
  mutate(
    model_type = "Neural Network",
    variable_selection = "Random Forest Importance",
    sample = "Undersampled",
    train_control = "10-fold cross-validation",
    size = nn_rf_us$bestTune$size,
    decay = nn_rf_us$bestTune$decay
  )

saveRDS(nn_rf_us, file.path(project_dir, "nn_rf_us_model.rds"))
saveRDS(nn_rf_us_results, file.path(project_dir, "nn_rf_us_results.rds"))
```

# **Neural Network Model Comparison**

```{r}
nn_results <- bind_rows(
  nn_logistic_results,
  nn_logistic_us_results,
  nn_rf_results,
  nn_rf_us_results
) %>%
  arrange(desc(f1_score))

nn_results %>%
  knitr::kable(
    digits = 3,
    caption = "Neural network model comparison ordered by F1 score."
  )
```

# **Save Final Neural Network Results**

```{r}
saveRDS(nn_results, file.path(project_dir, "nn_results.rds"))
```