---
title: "Random Forest Ensemble Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Introduction**

Decision trees are widely used supervised learning models for classification and regression because they provide transparent, rule-based predictions. However, individual decision trees are often considered weak learners because their predictive performance can be unstable across different training samples. A single tree may overfit the training data, become sensitive to small changes in the dataset, and produce predictions with high variance. These limitations are especially important in loan default prediction, where the model must generalize to unseen borrower records rather than simply memorize patterns in the training dataset.

Ensemble methods address these limitations by combining predictions from multiple models. Instead of relying on one decision tree, an ensemble fits many trees and aggregates their predictions. For classification tasks, the final prediction is commonly determined by majority vote or by averaging predicted class probabilities. This aggregation reduces variance and can improve predictive performance, particularly when the individual trees are not highly correlated with one another. Two common tree-based ensemble methods are random forests and boosting. Random forests build many trees independently, whereas boosting builds trees sequentially so that later trees focus on observations that earlier trees predicted poorly.

Random forests use bootstrap aggregation, or bagging, to create a diverse collection of decision trees. Each tree is trained on a random sample of the training data drawn with replacement, meaning some observations may appear multiple times while others may be excluded from a given tree. In addition, random forests introduce randomness at each split by allowing each tree to consider only a subset of predictors rather than all available predictors. This random feature selection reduces correlation among trees and helps prevent any single strong predictor from dominating every split. For classification, each tree votes for a class, and the forest combines these votes to produce the final prediction.

In this project, random forest models were evaluated for binary classification of loan default. The response variable, `loan_default`, indicates whether a borrower defaulted on a loan. Random forest models were trained on both the original imbalanced training dataset and an undersampled training dataset to assess the effect of class imbalance on default prediction. Models were also compared with and without cross-validation to evaluate whether tuning the number of predictors considered at each split improved predictive performance. Model performance was assessed on the holdout dataset using F1 score, accuracy, precision, recall, balanced accuracy, and area under the receiver operating characteristic curve. This approach allowed random forests to be compared with individual decision trees and other machine learning models while evaluating whether ensemble learning improved the identification of loan defaults.

# **Read in the Data**

## *Training Dataset*

```{r}
train <- readRDS("train.rds")
```

## *Undersampled Training Dataset*

```{r}
train_under <- readRDS("train_under.rds")
```

## *Holdout Dataset*

```{r}
hold <- readRDS("holdout.rds")
```

# **Preview the Data**

## *Training Dataset*

```{r}
skimr::skim(train)
```

## *Undersampled Training Dataset*

```{r}
skimr::skim(train_under)
```

## *Holdout Dataset*

```{r}
skimr::skim(hold)
```

# *Create `evaluate_classifier()`*

```{r}
evaluate_classifier <- function(actual, predicted, probability_yes, positive_class = "Yes") {
  actual <- factor(actual, levels = c("Yes", "No"))
  predicted <- factor(predicted, levels = c("Yes", "No"))

  cm <- caret::confusionMatrix(
    data = predicted,
    reference = actual,
    positive = positive_class
  )

  roc_obj <- pROC::roc(
    response = actual,
    predictor = probability_yes,
    levels = c("No", "Yes"),
    direction = "<",
    quiet = TRUE
  )

  precision <- cm$byClass["Precision"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall / (precision + recall)
  )

  tibble::tibble(
    accuracy = as.numeric(cm$overall["Accuracy"]),
    balanced_accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    precision = as.numeric(precision),
    sensitivity_recall = as.numeric(recall),
    specificity = as.numeric(cm$byClass["Specificity"]),
    f1_score = as.numeric(f1),
    auc = as.numeric(pROC::auc(roc_obj))
  )
}
```

# **`randomForest` Random Forest Model Without Cross Validation** 

## *Training Dataset*

### Build Ensemble & Train Model
____________________________________________________________________________________________________________


```{r}
# set.seed(123)
# 
# rf <- randomForest::randomForest(loan_default ~ .,                   
#                                  data=train,
#                                  ntree=30,
#                                  mtry=10,
#                                  nodesize=5,
#                                  importance=TRUE
#                                  )

rf <- readRDS("rf_model.rds")
```

### View Summary
____________________________________________________________________________________________________________

```{r}
rf
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
rf_prob <- predict(
  object = rf,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
rf_class <- predict(
  object = rf,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
rf_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(rf_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(rf_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(rf_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
rf_metrics <- evaluate_classifier(
  actual = rf_predictions$actual,
  predicted = rf_predictions$predicted,
  probability_yes = rf_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(rf, file.path(project_dir, "rf_model.rds"))
```

#### Save Results

```{r}
rf_results <- rf_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "randomForest",
    cross_validation = "No",
    ntree = "30",
    mtry = "10",
    nodesize = "5",
    sample = "Full"
  )

saveRDS(rf_results, file.path(project_dir, "rf_results.rds"))
```

## *Undersampled Training Dataset*

### Build Ensemble & Train Model
____________________________________________________________________________________________________________


```{r}
# set.seed(123)
# 
# rf_us <- randomForest::randomForest(loan_default ~ .,                   
#                                  data=train_under,
#                                  ntree=30,
#                                  mtry=10,
#                                  nodesize=5,
#                                  importance=TRUE
#                                  )

rf_us <- readRDS("rf_us_model.rds")
```

### View Summary
____________________________________________________________________________________________________________

```{r}
rf_us
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
rf_us_prob <- predict(
  object = rf_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
rf_us_class <- predict(
  object = rf_us,
  newdata = hold,
  type = "class"
)
```

#### Tabulate Predictions

```{r}
rf_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(rf_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(rf_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(rf_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
rf_us_metrics <- evaluate_classifier(
  actual = rf_us_predictions$actual,
  predicted = rf_us_predictions$predicted,
  probability_yes = rf_us_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(rf, file.path(project_dir, "rf_model.rds"))
```

#### Save Results

```{r}
rf_us_results <- rf_us_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "randomForest",
    cross_validation = "No",
    ntree = "30",
    mtry = "10",
    nodesize = "5",
    sample = "Undersampled"
  )

saveRDS(rf_us_results, file.path(project_dir, "rf_us_results.rds"))
```


# **`randomForest` Random Forest Model With Cross Validation**

## *Training Dataset*

### Detect Cores
____________________________________________________________________________________________________________

```{r}
# cores=parallel::detectCores()
# cores
```

### Set Up CPU Cores for Parallel Execution
____________________________________________________________________________________________________________

```{r}
# cl <- parallel::makeCluster(10)
```

### Register Parallel Backend
____________________________________________________________________________________________________________

```{r}
# registerDoParallel(cl)
```

### Set Up `trainControl()`
____________________________________________________________________________________________________________

```{r}
# cv_control <- trainControl(method = "cv", number = 5)
```

### Define Grid
____________________________________________________________________________________________________________

```{r}
# mtry_grid <- expand.grid(mtry = seq(1, ncol(train) - 1, by = 1))
```

### Build Ensemble & Train Model
____________________________________________________________________________________________________________

```{r}
# set.seed(123)
# 
# rf_cv <- train(loan_default ~ .,
#                   data = train,
#                   method = "rf",
#                   trControl = cv_control,
#                   tuneGrid = mtry_grid,
#                   ntree = 30,
#                   importance = TRUE,
#                   nodesize = 5)

rf_cv <- readRDS("rf_cv_model.rds")
```

### Shut Down Parallel Cluster
____________________________________________________________________________________________________________

```{r}
# stopCluster(cl) 
```

### Reset to Sequential Processing
____________________________________________________________________________________________________________

```{r}
# registerDoSEQ()
```

### View Summary
____________________________________________________________________________________________________________

```{r}
rf_cv
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
rf_cv_prob <- predict(
  object = rf_cv,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
rf_cv_class <- predict(
  object = rf_cv,
  newdata = hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
rf_cv_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(rf_cv_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(rf_cv_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(rf_cv_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
rf_cv_metrics <- evaluate_classifier(
  actual = rf_cv_predictions$actual,
  predicted = rf_cv_predictions$predicted,
  probability_yes = rf_cv_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(rf_cv, file.path(project_dir, "rf_cv_model.rds"))
```

#### Save Results

```{r}
rf_cv_results <- rf_cv_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "randomForest",
    cross_validation = "Yes",
    ntree = "30",
    mtry = "10",
    nodesize = "5",
    sample = "Full"
  )

saveRDS(rf_cv_results, file.path(project_dir, "rf_cv_results.rds"))
```

## *Undersampled Training Dataset*

### Detect Cores
____________________________________________________________________________________________________________

```{r}
# cores=parallel::detectCores()
# cores
```

### Set Up CPU Cores for Parallel Execution
____________________________________________________________________________________________________________

```{r}
# cl <- parallel::makeCluster(10)
```

### Register Parallel Backend
____________________________________________________________________________________________________________

```{r}
# registerDoParallel(cl)
```

### Set Up `trainControl()`
____________________________________________________________________________________________________________

```{r}
# cv_us_control <- trainControl(method = "cv", number = 5)
```

### Define Grid
____________________________________________________________________________________________________________

```{r}
# mtry_us_grid <- expand.grid(mtry = seq(1, ncol(train) - 1, by = 1))
```

### Build Ensemble & Train Model
____________________________________________________________________________________________________________

```{r}
# set.seed(123)
# 
# rf_cv_us <- train(loan_default ~ .,
#                   data = train_under,
#                   method = "rf",
#                   trControl = cv_us_control,
#                   tuneGrid = mtry_us_grid,
#                   ntree = 30,
#                   importance = TRUE,
#                   nodesize = 5)

rf_cv_us <- readRDS("rf_cv_us_model.rds")

```

### Shut Down Parallel Cluster
____________________________________________________________________________________________________________

```{r}
# stopCluster(cl) 
```

### Reset to Sequential Processing
____________________________________________________________________________________________________________

```{r}
# registerDoSEQ()
```

### View Summary
____________________________________________________________________________________________________________

```{r}
rf_cv_us
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
rf_cv_us_prob <- predict(
  object = rf_cv_us,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
rf_cv_us_class <- predict(
  object = rf_cv_us,
  newdata = hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
rf_cv_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(rf_cv_us_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(rf_cv_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(rf_cv_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
rf_cv_us_metrics <- evaluate_classifier(
  actual = rf_cv_us_predictions$actual,
  predicted = rf_cv_us_predictions$predicted,
  probability_yes = rf_cv_us_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(rf_cv_us, file.path(project_dir, "rf_cv_us_model.rds"))
```

#### Save Results

```{r}
rf_cv_us_results <- rf_cv_us_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "randomForest",
    cross_validation = "Yes",
    ntree = "30",
    mtry = "10",
    nodesize = "5",
    sample = "Undersampled"
  )

saveRDS(rf_cv_us_results, file.path(project_dir, "rf_cv_us_results.rds"))
```

# **`ranger` Random Forest Model Without Cross Validation**

## *Training Dataset*

### Build Ensemble & Train Model
____________________________________________________________________________________________________________

```{r}
# set.seed(123)
# 
# ranger_rf <- ranger::ranger(
#   formula = loan_default ~ .,
#   data = train,
#   classification = TRUE,
#   probability = TRUE,
#   num.trees = 30,
#   mtry = 10,
#   min.node.size = 5,
#   splitrule = "gini",
#   importance = "impurity",
#   seed = 123
# )

ranger_rf <- readRDS("ranger_rf_model.rds")
```

### View Summary
____________________________________________________________________________________________________________

```{r}
ranger_rf
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
ranger_rf_prob <- predict(
  object = ranger_rf,
  data = hold
)$predictions
```

#### Predict Class

```{r}
ranger_rf_class <- factor(
  ifelse(ranger_rf_prob[, "Yes"] >= 0.50, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
ranger_rf_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = ranger_rf_class,
  probability_yes = as.numeric(ranger_rf_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(ranger_rf_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
ranger_rf_metrics <- evaluate_classifier(
  actual = ranger_rf_predictions$actual,
  predicted = ranger_rf_predictions$predicted,
  probability_yes = ranger_rf_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(ranger_rf, file.path(project_dir, "ranger_rf_model.rds"))
```

#### Save Results

```{r}
ranger_rf_results <- ranger_rf_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "ranger",
    cross_validation = "No",
    num_trees = 30,
    mtry = 10,
    splitrule = "gini",
    min_node_size = 5,
    sample = "Full"
  )

saveRDS(ranger_rf_results, file.path(project_dir, "ranger_rf_results.rds"))
```

## *Undersampled Training Dataset*

### Build Ensemble & Train Model
____________________________________________________________________________________________________________

```{r}
# set.seed(123)
# 
# ranger_rf_us <- ranger::ranger(
#   formula = loan_default ~ .,
#   data = train_under,
#   classification = TRUE,
#   probability = TRUE,
#   num.trees = 30,
#   mtry = 10,
#   min.node.size = 5,
#   splitrule = "gini",
#   importance = "impurity",
#   seed = 123
# )

ranger_rf_us <- readRDS("ranger_rf_us_model.rds")
```

### View Summary
____________________________________________________________________________________________________________

```{r}
ranger_rf_us
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
ranger_rf_us_prob <- predict(
  object = ranger_rf_us,
  data = hold
)$predictions
```

#### Predict Class

```{r}
ranger_rf_us_class <- factor(
  ifelse(ranger_rf_us_prob[, "Yes"] >= 0.50, "Yes", "No"),
  levels = c("Yes", "No")
)
```

#### Tabulate Predictions

```{r}
ranger_rf_us_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("Yes", "No")),
  predicted = ranger_rf_us_class,
  probability_yes = as.numeric(ranger_rf_us_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(ranger_rf_us_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
ranger_rf_us_metrics <- evaluate_classifier(
  actual = ranger_rf_us_predictions$actual,
  predicted = ranger_rf_us_predictions$predicted,
  probability_yes = ranger_rf_us_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(ranger_rf_us, file.path(project_dir, "ranger_rf_us_model.rds"))
```

#### Save Results

```{r}
ranger_rf_us_results <- ranger_rf_us_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "ranger",
    cross_validation = "No",
    num_trees = 30,
    mtry = 10,
    splitrule = "gini",
    min_node_size = 5,
    sample = "Undersampled"
  )

saveRDS(ranger_rf_us_results, file.path(project_dir, "ranger_rf_us_results.rds"))
```

# **`ranger` Random Forest Model With Cross Validation**

## *Training Dataset*

### Detect Cores
____________________________________________________________________________________________________________

```{r}
# cores <- parallel::detectCores()
# cores
```

### Set Up CPU Cores for Parallel Execution
____________________________________________________________________________________________________________

```{r}
# cl <- parallel::makeCluster(10)
```

### Register Parallel Backend
____________________________________________________________________________________________________________

```{r}
# doParallel::registerDoParallel(cl)
```

### Set Up `trainControl()`
____________________________________________________________________________________________________________

```{r}
# ranger_cv_control <- caret::trainControl(
#   method = "cv",
#   number = 5,
#   classProbs = TRUE,
#   summaryFunction = caret::twoClassSummary,
#   savePredictions = "final",
#   allowParallel = TRUE
# )
```

### Define Grid
____________________________________________________________________________________________________________

```{r}
# ranger_grid <- expand.grid(
#   mtry = seq(1, ncol(train) - 1, by = 1),
#   splitrule = "gini",
#   min.node.size = 5
# )
```

### Build Ensemble & Train Model
____________________________________________________________________________________________________________

```{r}
# set.seed(123)
# 
# ranger_cv <- caret::train(
#   loan_default ~ .,
#   data = train,
#   method = "ranger",
#   trControl = ranger_cv_control,
#   tuneGrid = ranger_grid,
#   metric = "ROC",
#   num.trees = 30,
#   importance = "impurity"
# )

ranger_cv <- readRDS("ranger_cv_model.rds")
```

### Shut Down Parallel Cluster
____________________________________________________________________________________________________________

```{r}
# parallel::stopCluster(c1)
```

### Reset to Sequential Processing
____________________________________________________________________________________________________________

```{r}
# doParallel::registerDoSEQ()
```

### View Summary
____________________________________________________________________________________________________________

```{r}
ranger_cv
```

```{r}
ranger_cv$bestTune
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________

#### Predict Probabilities

```{r}
ranger_cv_prob <- predict(
  object = ranger_cv,
  newdata = hold,
  type = "prob"
)
```

#### Predict Class

```{r}
ranger_cv_class <- predict(
  object = ranger_cv,
  newdata = hold,
  type = "raw"
)
```

#### Tabulate Predictions

```{r}
ranger_cv_predictions <- tibble::tibble(
  actual = factor(hold$loan_default, levels = c("No", "Yes")),
  predicted = factor(ranger_cv_class, levels = c("No", "Yes")),
  probability_yes = as.numeric(ranger_cv_prob[, "Yes"])
)
```

#### Preview Table

```{r}
head(ranger_cv_predictions)
```

### Evaluate
____________________________________________________________________________________________________________

```{r}
ranger_cv_metrics <- evaluate_classifier(
  actual = ranger_cv_predictions$actual,
  predicted = ranger_cv_predictions$predicted,
  probability_yes = ranger_cv_predictions$probability_yes
)
```

### Store & Export
____________________________________________________________________________________________________________

#### Save Model

```{r}
saveRDS(ranger_cv, file.path(project_dir, "ranger_cv_model.rds"))
```

#### Save Results

```{r}
ranger_cv_results <- ranger_cv_metrics %>%
  mutate(
    model_type = "Random Forest Ensemble",
    package = "ranger",
    cross_validation = "Yes",
    ntree = "30",
    mtry = "10",
    nodesize = "5",
    sample = "Full"
  )

saveRDS(ranger_cv_results, file.path(project_dir, "ranger_cv_results.rds"))
```

## *Undersampled Training Dataset*

## Detect Cores
____________________________________________________________________________________________________________

```{r}
cores <- parallel::detectCores()
cores
```

### Set Up CPU Cores for Parallel Execution
____________________________________________________________________________________________________________

```{r}
cl <- parallel::makeCluster(10)
```

### Register Parallel Backend
____________________________________________________________________________________________________________

```{r}
doParallel::registerDoParallel(cl)
```

### Set Up `trainControl()`
____________________________________________________________________________________________________________

```{r}
ranger_cv_us_control <- caret::trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)
```

### Define Grid
____________________________________________________________________________________________________________

```{r}
ranger_us_grid <- expand.grid(
  mtry = seq(1, ncol(train_under) - 1, by = 1),
  splitrule = "gini",
  min.node.size = 5
)
```